### 2015-04-15 paper.tex
first commit

\documentclass{llncs}
\usepackage[utf8]{inputenc}
\usepackage[noend]{algorithmic}
\usepackage{algorithm}
\usepackage{epsfig}
\usepackage{multicol}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{subfigure}
\usepackage{url}
\newcommand{\verbatimproperties}{\renewcommand{\baselinestretch}{0.85} \small}
%\newcommand{\tableproperties}{\centering \big}
\newcommand{\LINECOMMENT}[1]{$\triangleright$ #1}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\title{On Extending a Full-Sharing Design with Batched Scheduling}

\author{Miguel Areias \and Ricardo Rocha}

\institute{CRACS \& INESC TEC and Faculty of Sciences, University of Porto\\
           Rua do Campo Alegre, 1021, 4169-007 Porto, Portugal\\
           \email{\{miguel-areias,ricroc\}@dcc.fc.up.pt}}

\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstract}

  \textbf{Keywords:} Multithreading, Tabling, Concurrency, Batched
  Scheduling
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Background}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{No-Sharing, Subgoal-Sharing and Full-Sharing Designs}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Batched Scheduling}

The decision about the evaluation flow is determined by the
\emph{scheduling strategy}. Different strategies may have a
significant impact on performance, and may lead to a different
ordering of solutions to the query goal. Arguably, the two most
successful tabling scheduling strategies are \emph{local scheduling}
and \emph{batched scheduling}~\cite{Freire-96}.

Local scheduling strategy schedules the evaluation of a program in a
breath-first manner. It favors the backtracking first with completion
instead of the forward execution, leaving the consumption of answers
for last. Thus, it only allows a Cluster of Dependent Subgoals (CDS)
to return answers only after the completion point has been
reached~\cite{Freire-96}. In other words, the local scheduling tries
to keep a CDS as minimal as possible. When new answers are found, they
are added to the table space and the computation fails as consequence,
tabled subgoals inside a CDS propagate their answers to outside the
CDS only after its completion point is found. Local scheduling causes
a sooner completion of subgoals, which creates less complex
dependencies between them.

On the other hand, batched scheduling schedules the evaluation of a
program in a depth-first manner. It favors the forward execution first
instead of backtracking, leaving the consumption of answers and
completion for last. It thus tries to delay the need to move around
the search tree by batching the return of answers. When new answers
are found for a particular tabled subgoal, they are added to the table
space and the execution continues. For some situations, this results
in creating dependencies to older subgoals, therefore enlarging the
current CDS~\cite{Sagonas-98} and delaying the completion point to an
older generator node.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Full-Sharing with Batched Scheduling}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Our Approach}

The key idea of our approach, which we named Privately-consumed Answer
Chaining (PAC), is then to extend the \emph{FS} design with batched
scheduling, by chaining privately for each subgoal call, the answers
that were already consumed by a thread. Since the procedure is
private, it will only affect the thread that is doing it. At the end,
when the evaluation is complete, i.e, when a subgoal call is marked as
complete, we put one of the private chain as public, so that from that
point on all threads can use that chain in complete (only reading)
mode.

Figure~\ref{fig_tabtries_pcc} shows the key data structures for
supporting the implementation of the PAC procedure during the
evaluation of a tabled subgoal call $P_{i.j}$ using the FS design. The
FS design, uses a subgoal entry data structure to store common
information for a subgoal call and a subgoal frame (\emph{SF}) data
structure to store private information about the execution of each
thread. The PAC procedure works at the subgoal frame level, which is
private to each thread.

\begin{figure}[!ht]
\centering
\includegraphics[width=10.5cm]{figures/pcc.pdf}
\caption{The \emph{FS} design with the PAC procedure - (a) private
  chaining and (b) public chaining}
\label{fig_tabtries_pcc}
\end{figure}

Figure~\ref{fig_tabtries_pcc}(a) shows then a situation where two
threads, $T_1$ and $T_{k-1}$, are sharing the same subgoal entry call
$P_{i.j}$ when the subgoal is still under evaluation, i.e., the
subgoal is not yet complete. The current state of the evaluation shows
an answer trie with $3$ answers found for the subgoal call
$P_{i.j}$. For the sake of simplicity, we are omitting the internal
answer trie nodes and we are only showing the leaf nodes nodes $LN_1$,
$LN_2$ and $LN_3$, in the figure. With the PAC procedure, the leaf
nodes are not chained in the \emph{AT} data structure. Now, the
chaining process is done privately, and for that, we use the subgoal
frame structure of each thread. On the subgoal frame structure we
added a new field, called \emph{consumer}, to store the answers found
within the execution of the thread. In order to minimize the impact of
the \emph{PAC} optimization, each node within the new consumer answer
structure has two fields: (i) an entry pointer, which points to the
corresponding leaf node in the answer trie data structure; (ii) a next
pointer to chain the answers within the consumer structure. To
maintain a good performance, when the number of nodes exceeds a
certain threshold, we use a hash trie mechanism design similar to the
one presented in the work~\cite{Areias-ijpp15}. However, since this
mechanism is private to each thread, it does not require any of the
tools that were necessary to support concurrency. In particular, on
each hash trie level, we have removed the tools necessary to support
concurrency, such as useless pointers and compare-and-swap
operations. We have chosen this hashing mechanism, because it showed a
good balance between lookup and insert
operations~\cite{Areias-ijpp15}, but the major reason was mostly
because of the integration in the TabMalloc memory
allocator~\cite{Areias-12b}.

Going back to Figure~\ref{fig_tabtries_pcc}(a), the consumer answer
structures represent then two different situations where threads can
be evaluating a subgoal call. Thread $T_1$ has only found one answer
and it is using a direct consumer answer chaining to access the node
$LN_1$. Thread $T_{k-1}$ was already found three answers for the
subgoal call and it is already using the hash trie mechanism within
its consumer answer structure. The consumer nodes are chained between
themselves, thus that consumer nodes belonging to thread $T_{k-1}$ can
consume the answers as in the original mechanism.

Figure~\ref{fig_tabtries_pcc}(b) shows the state of the subgoal call
after completion (recall that after completion of a subgoal call, the
threads use loader nodes to consume the answers). When a thread $T$
completes a subgoal call, it frees its private consumer structures,
but before doing that, it checks whether another thread as already
marked the subgoal as completed. If no other thread has done that,
then thread $T$ not only follows its private chaining mechanism as it
would for freeing its private nodes, but also, follows the pointers to
the answer trie leaf nodes in order to reproduce the chain inside the
answer trie. Since this procedure is done inside a critical region, no
more than one thread can be doing this chaining process. Thus, in
Figure~\ref{fig_tabtries_pcc}(b), we are showing a situation where the
subgoal call is completed and both threads $T_1$ and $T_{k-1}$ have
already removed their consumer answer structures and chained the leaf
nodes inside the answer trie.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Implementations details}

At the implementation level, the major difference between local and
batched scheduling is in the tabling operation \emph{tabled new
  answer}, where we decide what to do when an answer is found during
the evaluation. This operation checks whether a newly found answer is
already in the corresponding answer trie structure and, if not,
inserts it. For the NS and SS designs the support for batched
scheduling is immediate, since the answer trie data structure is not
shared among threads. The usage of batched scheduling with the FS
design requires further support since with batched scheduling, answers
are immediately propagated and we have to ensure that the propagation
of an answer occurs on all subgoal calls one and only once. To do so,
we take advantage of the private chaining procedure, presented in the
previous subsection, as a way to keep, for every subgoal call of every
thread, track of all the answers that were already propagated. This
requires minor changes to the \emph{tabled new answer} tabling
operation. Algorithm~\ref{alg_table_new_answer_batched} shows how we
have extended the tabled new answer operation to support the FS design
with batched scheduling.

\begin{algorithm} [ht]
\caption{tabled\_new\_answer(answer ANS, subgoal frame SF)}
\algsetup{indent=0.65cm}
\begin{algorithmic}[1]
  \STATE $leaf \gets check\_insert\_answer\_trie(ANS, SF)$
  \IF {$NS\_design\ or\ SS\_design$}
    \STATE $. . .$                         \COMMENT{without changes}
  \ELSE                                                  [FS design]
    \STATE $chain \gets check\_insert\_consumer\_chain(leaf, SF)$
    \IF {$is\_answer\_marked\_as\_found(chain) = True$}
      \RETURN $failure$
     \ELSE                                        [the answer is new]
       \STATE $mark\_answer\_as\_found(chain)$
       \IF {$local\_scheduling\_mode(SF)$}
         \RETURN $failure$
       \ELSE                                    [batched scheduling mode]
         \RETURN $proceed$
       \ENDIF
     \ENDIF
   \ENDIF
\end{algorithmic}
\label{alg_table_new_answer_batched}
\end{algorithm}

The algorithm receives two arguments: the new answer found during the
evaluation (\emph{ANS}) and the subgoal frame which corresponds to the
call at hand (\emph{SF}). The \emph{NS\_design}, \emph{SS\_design} and
\emph{FS\_design} macros define which table design is enabled.

The algorithm begins by checking/inserting the given \emph{ANS} into
the answer trie structure, which will return the leaf node for the
path representing \emph{ANS} (line 1). In line 2, it then tests
whether one of the NS or SS designs are active, and in such a case,
the algorithm is remains unchanged.

Otherwise, for the FS design (lines 4 to 13), it checks/inserts
the given \emph{leaf} node into the private consumer chain for the
current thread, which will return the corresponding chain node. In
line 6, it then tests whether the chain node already existed in the
consumer chain, i.e., if it was inserted or not by the current
check/insert operation in order to return failure (line 7), or it
proceed with marking the answer \emph{ANS} has found (line 9). At the
end (lines 10 to 13), it returns failure if local scheduling is active
(line 11), otherwise, the batched scheduling is active, thus it
propagates the answer \emph{ANS} (line 13).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Performance Analysis}

We now present experimental results about the usage of the batched
scheduling on the \emph{NS}, \emph{SS} and \emph{FS} designs. For the
sake of simplicity, for the \emph{SS} and \emph{FS} designs, we will be
presenting only the results for the lock free \emph{LF2} proposal
(\emph{SS}$_{LF_2}$ and \emph{FS}$_{LF_2}$), since they were the ones
that presented the lowest overheads in the previous chapters. For the
\emph{FS}$_{LF_2}$ design, we will use it with the \emph{PCC} procedure
enabled.

Concerning the benchmarks, we will be using the same five sets of
benchmarks presented before with the same number of runs per
benchmark, the same formula to calculate the overhead ratios, and the
same worst case scenario approach, where all threads begin with the
same query goal. To put the results in perspective, we experimented
with $1$, $8$, $16$, $24$ and $32$ threads (the maximum number of
cores available in our machine) with batched and local scheduling.

Table~\ref{tab_batched_overhead} shows the overhead ratios, when
compared with the \emph{NS} design with 1 thread (running with local
scheduling, PtMalloc and without TabMalloc), for the \emph{NS},
\emph{SS}$_{LF_2}$ and \emph{FS}$_{LF_2+PCC}$ designs (all running with
TabMalloc and TcMalloc), when running 1, 8, 16, 24 and 32 threads with
local and batched scheduling on the five sets of benchmarks (the
results by set of benchmark can be seen in the
Appendix.~\ref{app:batched}). For each design, the table has then two
columns, a column with \textbf{Local} that shows results already
presented in previous chapters for the local scheduling and a column
with \textbf{Batched} that shows the new results with the batched
scheduling. The overhead results presented in both \textbf{Local} and
\textbf{Batched} columns use as base time the execution times
presented in the \emph{NS} column of the Table~\ref{tab_benchs}.

\begin{table}[!ht]
\centering
\caption{Overhead ratios, when compared with the NS design with $1$
  thread (running with local scheduling, PtMalloc and without
  TabMalloc) for the NS, SS$_{LF_2}$, FS$_{LF_2+PCC}$ designs (with
  TabMalloc and TcMalloc), when running 1, 8, 16, 24 and 32 threads
  with local and batched scheduling on the five sets of benchmarks
  (best ratios by row and by design for the Minimum, Average and
  Maximum are in bold)}
%\scalebox{1.00} {
\begin{tabular}{ll|rr|rr|rr}
\hline\hline
\multicolumn{2}{c|}{\multirow{2}{*}{\bf Threads}} &
\multicolumn{2}{c|}{\multirow{1}{*}{\bf NS}} &
\multicolumn{2}{c}{\multirow{1}{*}{\bf SS$_{LF_2}$}} & 
\multicolumn{2}{|c}{\multirow{1}{*}{\bf FS$_{LF_2+PCC}$}}\\ \cline{3-8}
& 
& \multicolumn{1}{c}{\bf Local}
& \multicolumn{1}{c}{\bf Batched}
& \multicolumn{1}{|c}{\bf Local}
& \multicolumn{1}{c}{\bf Batched}
& \multicolumn{1}{|c}{\bf Local}
& \multicolumn{1}{c}{\bf Batched}\\
\hline
\multirow{4}{*}{\bf 1}
& {\bf Min }& {\bf 0.53}& 0.55& {\bf 0.54}& 0.55& 1.01& {\bf 0.95}\\
& {\bf Avg }& {\bf 0.78}& 0.82& {\bf 0.84}& 0.90& {\bf 1.30}& 1.46\\
& {\bf Max }& 1.06& {\bf 1.05}& {\bf 1.04}& {\bf 1.04}& {\bf 1.76}& 2.33\\
& {\bf StD }& 0.15& 0.14& 0.17& 0.16& 0.22& 0.44\\
\hline
\multirow{4}{*}{\bf 8}
& {\bf Min }& 0.66& {\bf 0.63}& 0.66& {\bf 0.63}& 1.16&{\bf  0.99}\\
& {\bf Avg }& {\bf 0.85}& 0.88& {\bf 0.92}& 0.93& {\bf 1.88}& 1.95\\
& {\bf Max }& {\bf 1.12}& 1.14& 1.20& {\bf 1.15}& {\bf 2.82}& 3.49\\
& {\bf StD }& 0.13& 0.14& 0.15& 0.14& 0.60& 0.79\\
\hline
\multirow{4}{*}{\bf 16}
& {\bf Min }& 0.85& {\bf 0.75}& 0.82& {\bf 0.77}& 1.17& {\bf 1.06}\\
& {\bf Avg }& {\bf 0.98}& 1.00& {\bf 1.04}& 1.05& {\bf 1.97}& 2.08\\
& {\bf Max }& {\bf 1.16}& 1.31& 1.31& {\bf 1.28}& {\bf 3.14}& 3.69\\
& {\bf StD }& 0.09& 0.17& 0.12& 0.13& 0.65& 0.83\\
\hline
\multirow{4}{*}{\bf 24}
& {\bf Min }& {\bf 0.91}& 0.93& 1.02& {\bf 0.98}& 1.16& {\bf 1.09}\\
& {\bf Avg }& {\bf 1.15}& 1.16& 1.22& {\bf 1.19}& {\bf 2.06}& 2.19\\
& {\bf Max }& 1.72& {\bf 1.60}& 1.81& {\bf 1.61}& {\bf 3.49}& 4.08\\
& {\bf StD }& 0.20& 0.21& 0.18& 0.16& 0.70& 0.91\\
\hline
\multirow{4}{*}{\bf 32}
& {\bf Min }& 1.05& {\bf 1.04}& {\bf 1.07}& 1.12& 1.33& {\bf 1.26}\\
& {\bf Avg }& 1.51& {\bf 1.49}& 1.54& {\bf 1.51}& {\bf 2.24}& 2.41\\
& {\bf Max }& {\bf 2.52}& 2.63& {\bf 2.52}& 2.62& {\bf 3.71}& 4.51\\
& {\bf StD }& 0.45& 0.45& 0.42& 0.43& 0.74& 1.02\\
\hline\hline
\end{tabular}%}
\label{tab_batched_overhead}
\end{table}

By observing Table~\ref{tab_batched_overhead}, we can see that, for
one thread, on average, local scheduling is sightly better than
batched on the three designs. For the \emph{NS} design we have $0.78$
and $0.82$, for the \emph{SS}$_{LF_2}$ design we have $0.84$ and $0.90$
and for the \emph{FS}$_{LF_2+PCC}$ design we have $1.30$ and $1.46$
average overhead ratios, for the local and batched scheduling
strategies, respectively.

As we scale the number of threads, one can observe that, for the
\emph{NS} and \emph{SS}$_{LF_2}$ designs both scheduling strategies have
similar minimum, average and maximum overhead ratios. For the
\emph{FS}$_{LF_2+PCC}$ design, the best minimum overhead ratio is
always for batched scheduling. The reader can observe on
Appendix~\ref{app:batched} that the minimum overhead values for $8$,
$16$, $24$ and $32$ threads are given by the benchmark belonging to
the model checking set (see Table~\ref{tab_benchs} for the
characteristics of the benchmarks). For the average and maximum
overhead ratio, local scheduling is always better than batched
scheduling. The reader can observe on Appendix.~\ref{app:batched} that
the maximum overhead values for $8$, $16$, $24$ and $32$ threads are
given by the pyramid benchmark in the path right set.

In summary, we can say that both the local and batched scheduling
strategies have similar overhead results on worst case scenarios for
the \emph{NS}, \emph{SS}$_{LF_2}$ and \emph{FS}$_{LF_2+PCC}$ designs.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions and Further Work}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Acknowledgments}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bibliographystyle{splncs}
\bibliography{references}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

### 2015-04-15 paper.tex
performance on worst case scenarios

@@
 We now present experimental results about the usage of the batched
-scheduling on the \emph{NS}, \emph{SS} and \emph{FS} designs. For the
-sake of simplicity, for the \emph{SS} and \emph{FS} designs, we will be
-presenting only the results for the lock free \emph{LF2} proposal
-(\emph{SS}$_{LF_2}$ and \emph{FS}$_{LF_2}$), since they were the ones
-that presented the lowest overheads in the previous chapters. For the
-\emph{FS}$_{LF_2}$ design, we will use it with the \emph{PCC} procedure
-enabled.
+scheduling on the \emph{NS}, \emph{SS} and \emph{FS} designs. The
+environment for our experiments was a machine with 32-Core AMD Opteron
+(TM) Processor 6274 (2 sockets with 16 cores each) with 32G of main
+memory, each processor with caches $L1$, $L2$ and $L3$ respectively
+with the sizes of 64K, 2048K and 6144K, running the Linux kernel is
+the 3.16.7-200.fc20.x86\_64, with Yap 6.3 compiled with gcc 4.8.
 
-Concerning the benchmarks, we will be using the same five sets of
-benchmarks presented before with the same number of runs per
-benchmark, the same formula to calculate the overhead ratios, and the
-same worst case scenario approach, where all threads begin with the
-same query goal. To put the results in perspective, we experimented
-with $1$, $8$, $16$, $24$ and $32$ threads (the maximum number of
-cores available in our machine) with batched and local scheduling.
+We used five sets of benchmarks. The \emph{Large Joins} and
+\emph{WordNet} sets were obtained from the OpenRuleBench
+project~\cite{Liang-09}; the \emph{Model Checking} set includes three
+different specifications and transition relation graphs usually used
+in model checking applications; the \emph{Path Left} and \emph{Path
+  Right} sets implement two recursive definitions of the well-known
+$path/2$ predicate, that computes the transitive closure in a graph,
+using several different configurations of $edge/2$ facts. We
+experimented graphs with the \emph{BTree} configuration with depth 17,
+the \emph{Pyramid} and \emph{Cycle} configurations with depth 2000 and
+the \emph{Grid} configuration with depth 35.
+
+In order to have a deeper insight on the behavior of each benchmark,
+and therefore clarify some of the results that are presented next, we
+first characterize the benchmarks. The columns in
+Table~\ref{tab_benchs} have the following meaning: (i) {\bf calls:} is
+the number of different calls to tabled subgoals. It corresponds to
+the number of paths in the subgoal tries. (ii) {\bf trie nodes:} is
+the total number of trie nodes allocated in the corresponding
+subgoal/answer trie structures. (iii) {\bf trie depth:} is the
+minimum/average/maximum number of trie node levels required to
+represent a path in the corresponding subgoal/answer trie
+structures. Trie structures with smaller average values are more
+amenable to higher lock contention. (iv) {\bf unique:} is the number
+of different tabled answers found. It corresponds to the number of
+paths in the answer tries. (v) {\bf repeated:} is the number of
+redundant tabled answers found. (vi) {\bf NS:} is the average
+execution time, in seconds, of ten runs for 1 thread with the NS
+design without the TabMalloc memory allocator. In what follows, we
+will use these times as the base times when computing the overhead
+ratios for the other designs.
+
+By observing Table~\ref{tab_benchs}, the \emph{Mondial} benchmark,
+from the \emph{Large Joins} set, and the three \emph{Model
+  Checking} benchmarks seem to be the benchmarks least amenable to
+lock contention since they are the ones that find less unique answers
+and that have the deepest trie structures. In this regard, the
+\emph{Path Left} and \emph{Path Right} sets correspond to the
+opposite case. They find a huge number of answers and have very
+shallow trie structures. On the other hand, the \emph{WordNet} and
+\emph{Path Right} sets have the benchmarks with the largest number
+of different subgoal calls, which can reduce the probability of lock
+contention because answers can be found for different subgoal calls
+and therefore be inserted with minimum overlap. On the opposite side
+are the \emph{Join2} benchmark, from the \emph{Large Joins} set,
+and the \emph{Path Left} benchmarks, which have only a single tabled
+subgoal call.
+
+\begin{table}[ht]
+\centering
+\caption{Characteristics of the benchmark programs}
+\scalebox{0.82} { 
+\begin{tabular}{lrrccrrrcc}
+\hline\hline
+\multirow{2}{*}{\bf Bench}
+& \multicolumn{3}{c}{\bf Tabled Subgoals}
+&
+& \multicolumn{4}{c}{\bf Tabled Answers}
+& \multicolumn{1}{c}{\bf Time (sec)} \\ \cline{2-4}\cline{6-9}
+& \multicolumn{1}{c}{\bf calls}
+& \multicolumn{1}{c}{\bf trie nodes}
+& \multicolumn{1}{c}{\bf trie depth}
+&
+& \multicolumn{1}{c}{\bf unique}
+& \multicolumn{1}{c}{\bf repeated}
+& \multicolumn{1}{c}{\bf trie nodes}
+& \multicolumn{1}{c}{\bf trie depth}
+& \multicolumn{1}{c}{\bf NS} \\
+\hline
+\multicolumn{10}{l}{\bf Large Joins} \\
+\bf{Join2}    &       1 &       6 & 5/5/5 && 2,476,099 &         0 &
+2,613,660 &    5/5/5 &  2.85 \\
+\bf{Mondial}  &      35 &      42 & 3/4/4 &&     2,664 & 2,452,890 &
+   14,334 &    6/7/7 &  0.84 \\
+\hline
+\multicolumn{10}{l}{\bf WordNet} \\
+\bf{Clusters} & 117,659 & 235,319 & 2/2/2 &&   166,877 &   161,853 &
+284,536 &    1/1/1 &   0.83 \\
+\bf{Holo}     & 117,657 & 235,315 & 2/2/2 &&    74,838 &        54 &
+192,495 &    1/1/1 &   0.75 \\
+\bf{Hyper}    & 117,657 & 235,315 & 2/2/2 &&   698,472 &     8,658 &
+816,129 &    1/1/1 &   1.42 \\
+\bf{Hypo}     & 117,657 & 117,659 & 2/2/2 &&   698,472 &    20,341 &
+816,129 &    1/1/1 &   1.53 \\
+\bf{Mero}     & 117,657 & 117,659 & 2/2/2 &&    74,838 &        13 &
+192,495 &    1/1/1 &   0.74 \\
+\bf{Tropo}    & 117,657 & 235,315 & 2/2/2 &&       472 &         0 &
+118,129 &    1/1/1 &   0.66 \\
+\hline
+\multicolumn{10}{l}{\bf Model Checking} \\
+\bf{IProto}   &       1 &       6 & 5/5/5 &&   134,361 &   385,423 &
+1,554,896 &  4/51/67 & 2.70 \\
+\bf{Leader}   &       1 &       5 & 4/4/4 &&     1,728 &   574,786 &
+   41,788 & 15/80/97 & 3.51  \\
+\bf{Sieve}    &       1 &       7 & 6/6/6 &&       380 & 1,386,181 &
+    8,624 & 21/53/58 & 18.50 \\
+\hline
+\multicolumn{10}{l}{\bf Path Left} \\
+\bf{BTree}    &       1 &       3 & 2/2/2 && 1,966,082 &         0 &
+2,031,618 &    2/2/2 &   1.53 \\
+\bf{Cycle}    &       1 &       3 & 2/2/2 && 4,000,000 &     2,000 &
+4,002,001 &    2/2/2 &   3.52 \\
+\bf{Grid}     &       1 &       3 & 2/2/2 && 1,500,625 & 4,335,135 &
+1,501,851 &    2/2/2 &  1.93 \\
+\bf{Pyramid}  &       1 &       3 & 2/2/2 && 3,374,250 & 1,124,250 &
+3,377,250 &    2/2/2 &   3.08 \\
+\hline
+\multicolumn{10}{l}{\bf Path Right} \\
+\bf{BTree}    & 131,071 & 262,143 & 2/2/2 && 3,801,094 &         0 &
+3,997,700 &    1/2/2 &  2.33 \\
+\bf{Cycle}    &   2,001 &   4,003 & 2/2/2 && 8,000,000 &     4,000 &
+8,004,001 &    1/2/2 &  3.55 \\
+\bf{Grid}     &   1,226 &   2,453 & 2/2/2 && 3,001,250 & 8,670,270 &
+3,003,701 &    1/2/2 &  2.32 \\
+\bf{Pyramid}  &   3,000 &   6,001 & 2/2/2 && 6,745,501 & 2,247,001 &
+6,751,500 &    1/2/2 &  3.17 \\
+\hline\hline
+\end{tabular}}
+\label{tab_benchs}
+\end{table}
+
+Note that for the moment our goal is to evaluate the robustness of our
+implementation when exposed to \emph{worst case} scenarios. By
+focusing on \emph{worst case} scenarios, we show the lowest bounds in
+terms of performance that each design might achieve when applied/used
+with other real world applications/programs. Moreover, by testing the
+framework with worst case scenarios, we avoid the peculiarities of the
+program at hand and we try to focus on measuring the real value of our
+designs. Thus, we will follow a common approach to create worst case
+scenarios and we will run all threads starting with the same query
+goal. By doing this, it is expected that all threads will access the
+table space, to check/insert for subgoals and answers, at similar
+times, thus causing a huge stress on the same critical regions. To put
+the results in perspective, we experimented with $1$, $8$, $16$, $24$
+and $32$ threads (the maximum number of cores available in our
+machine) with local and batched scheduling on the five sets of
+benchmarks presented in the early subsection.
 
 Table~\ref{tab_batched_overhead} shows the overhead ratios, when
-compared with the \emph{NS} design with 1 thread (running with local
-scheduling, PtMalloc and without TabMalloc), for the \emph{NS},
-\emph{SS}$_{LF_2}$ and \emph{FS}$_{LF_2+PCC}$ designs (all running with
-TabMalloc and TcMalloc), when running 1, 8, 16, 24 and 32 threads with
-local and batched scheduling on the five sets of benchmarks (the
-results by set of benchmark can be seen in the
-Appendix.~\ref{app:batched}). For each design, the table has then two
-columns, a column with \textbf{Local} that shows results already
-presented in previous chapters for the local scheduling and a column
-with \textbf{Batched} that shows the new results with the batched
-scheduling. The overhead results presented in both \textbf{Local} and
-\textbf{Batched} columns use as base time the execution times
-presented in the \emph{NS} column of the Table~\ref{tab_benchs}.
+compared with the NS design with 1 thread (running with local
+scheduling), for the NS, SS and FS designs (all running TabMalloc),
+when running 1, 8, 16, 24 and 32 threads with local and batched
+scheduling on the five sets of benchmarks. For each design, the table
+has then two columns, a column with \emph{Local} that shows results
+for the local scheduling and a column with \emph{Batched} with results
+for batched scheduling. The overhead results presented in both
+\emph{Local} and \emph{Batched} columns use as base time the execution
+times presented in the \emph{NS} column of the Table~\ref{tab_benchs}.
 
@@
 \caption{Overhead ratios, when compared with the NS design with $1$
-  thread (running with local scheduling, PtMalloc and without
-  TabMalloc) for the NS, SS$_{LF_2}$, FS$_{LF_2+PCC}$ designs (with
-  TabMalloc and TcMalloc), when running 1, 8, 16, 24 and 32 threads
-  with local and batched scheduling on the five sets of benchmarks
-  (best ratios by row and by design for the Minimum, Average and
-  Maximum are in bold)}
+  thread (running with local scheduling without TabMalloc) for the NS,
+  SS, FS designs (with TabMalloc), when running 1, 8, 16, 24 and 32
+  threads with local and batched scheduling on the five sets of
+  benchmarks (best ratios by row and by design for the Minimum,
+  Average and Maximum are in bold)}
 %\scalebox{1.00} {
-\begin{tabular}{ll|rr|rr|rr}
+\begin{tabular}{ll|cc|cc|cc}
 \hline\hline
@@
 \multicolumn{2}{c|}{\multirow{1}{*}{\bf NS}} &
-\multicolumn{2}{c}{\multirow{1}{*}{\bf SS$_{LF_2}$}} & 
-\multicolumn{2}{|c}{\multirow{1}{*}{\bf FS$_{LF_2+PCC}$}}\\ \cline{3-8}
+\multicolumn{2}{c}{\multirow{1}{*}{\bf SS}} & 
+\multicolumn{2}{|c}{\multirow{1}{*}{\bf FS}}\\ \cline{3-8}
 & 
@@
 one thread, on average, local scheduling is sightly better than
-batched on the three designs. For the \emph{NS} design we have $0.78$
-and $0.82$, for the \emph{SS}$_{LF_2}$ design we have $0.84$ and $0.90$
-and for the \emph{FS}$_{LF_2+PCC}$ design we have $1.30$ and $1.46$
-average overhead ratios, for the local and batched scheduling
-strategies, respectively.
-
-As we scale the number of threads, one can observe that, for the
-\emph{NS} and \emph{SS}$_{LF_2}$ designs both scheduling strategies have
-similar minimum, average and maximum overhead ratios. For the
-\emph{FS}$_{LF_2+PCC}$ design, the best minimum overhead ratio is
-always for batched scheduling. The reader can observe on
-Appendix~\ref{app:batched} that the minimum overhead values for $8$,
-$16$, $24$ and $32$ threads are given by the benchmark belonging to
-the model checking set (see Table~\ref{tab_benchs} for the
-characteristics of the benchmarks). For the average and maximum
-overhead ratio, local scheduling is always better than batched
-scheduling. The reader can observe on Appendix.~\ref{app:batched} that
-the maximum overhead values for $8$, $16$, $24$ and $32$ threads are
-given by the pyramid benchmark in the path right set.
-
-In summary, we can say that both the local and batched scheduling
+batched on the three designs. For the NS design we have $0.78$ and
+$0.82$, for the SS design we have $0.84$ and $0.90$ and for the FS
+design we have $1.30$ and $1.46$ average overhead ratios, for the
+local and batched scheduling strategies, respectively. As we scale the
+number of threads, one can observe that, for the NS and SS designs
+both scheduling strategies have similar minimum, average and maximum
+overhead ratios. For the FS design, the best minimum overhead ratio is
+always for batched scheduling. For the average and maximum overhead
+ratio, local scheduling is always better than batched scheduling. In
+summary, we can say that both the local and batched scheduling
 strategies have similar overhead results on worst case scenarios for
-the \emph{NS}, \emph{SS}$_{LF_2}$ and \emph{FS}$_{LF_2+PCC}$ designs.
+the NS, SS and FS designs.
 
@@
 
-\section*{Acknowledgments}
+%\section*{Acknowledgments}
 

### 2015-04-15 paper.tex
performance on worst case scenarios

@@
 \newcommand{\LINECOMMENT}[1]{$\triangleright$ #1}
+
+%\setlength{\belowcaptionskip}{-26pt}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 scheduling on the five sets of benchmarks. For each design, the table
-has then two columns, a column with \emph{Local} that shows results
-for the local scheduling and a column with \emph{Batched} with results
-for batched scheduling. The overhead results presented in both
-\emph{Local} and \emph{Batched} columns use as base time the execution
-times presented in the \emph{NS} column of the Table~\ref{tab_benchs}.
+has then two columns, a column with {\bf Local} that shows results for
+the local scheduling and a column with {\bf Batched} with results for
+batched scheduling. The overhead results presented in both {\bf Local}
+and {\bf Batched} columns use as base time the execution times
+presented in the {\bf NS} column of the Table~\ref{tab_benchs}.
 

### 2015-04-15 paper.tex
introduction -> begin

@@
 
+Tabling~\cite{Chen-96} is an implementation technique that overcomes
+some limitations of traditional Prolog systems in dealing with
+redundant sub-computations and recursion. Tabling consists of storing
+intermediate answers for subgoals so that they can be reused when a
+repeated subgoal appears during the resolution process. Tabling has
+become a popular and successful technique thanks to the
+ground-breaking work in the XSB Prolog system and in particular in the
+SLG-WAM engine~\cite{Sagonas-98}, the most successful engine of
+XSB. The success of SLG-WAM led to several alternative implementations
+that differ in the execution rule, in the data-structures used to
+implement tabling, and in the changes to the underlying Prolog
+engine. Implementations of tabling are now widely available in systems
+like Yap Prolog, B-Prolog, ALS-Prolog, Mercury, Ciao Prolog and more
+recently Picat.
+
+Multithreading in Prolog is the ability to concurrently perform
+computations, in which each computation runs independently but shares
+the program clauses. When multithreading is combined with tabling, we
+have the best of both worlds, since we can exploit the combination of
+higher procedural control with higher declarative semantics. To the
+best of our knowledge, XSB~\cite{Marques-08} and Yap~\cite{Areias-12a}
+were the only Prolog systems that were able to combine both
+multithreading with tabling. Yap's threads have their own execution
+stacks and only share the code area where predicates, records, flags
+and other global non-backtrackable data are stored. For tabled
+evaluation, a thread views its tables as private but, at the engine
+level, Yap has three designs: (i) The {\bf No-Sharing (NS)} design,
+where each thread allocates fully private tables for each new tabled
+subgoal called during its computation. (ii) {\bf Subgoal-Sharing
+  (SS)}, where threads share part of the table space and the (iii)
+{\bf Full-Sharing (FS)}, where threads share the complete table
+space.
+
+In this paper we discuss a novel approach for solving the problem of
+supporting multithreaded batched scheduling in the FS design and we
+present a performance analysis comparison between local scheduling
+with batched scheduling. Experimental results show that despite the
+extra data structures required to extend the FS design with batched
+scheduling, the execution time of the FS design with batched
+scheduling is still quite competitive when comparing against local
+scheduling.
+
+The remainder of the paper is organized as follows. First, we briefly
+introduce some background and related work. Then, we describe our
+approach for combining batched scheduling with the FS design. Next, we
+show the most important implementation details and finally, we discuss
+experimental results and we end by outlining some conclusions.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Background}
+
+This section introduces some background needed for the following
+sections.
 

### 2015-04-15 paper.tex
introduction -> end

@@
 where each thread allocates fully private tables for each new tabled
-subgoal called during its computation. (ii) {\bf Subgoal-Sharing
-  (SS)}, where threads share part of the table space and the (iii)
-{\bf Full-Sharing (FS)}, where threads share the complete table
-space.
+subgoal called during its computation. (ii) The {\bf Subgoal-Sharing
+  (SS)}, where threads share part of the table space and the (iii) The
+{\bf Full-Sharing (FS)}, where threads share the complete table space.
 

### 2015-04-15 paper.tex
implementation details -> end

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Implementations details}
+\subsection{Implementations Details}
 
@@
 already in the corresponding answer trie structure and, if not,
-inserts it. For the NS and SS designs the support for batched
-scheduling is immediate, since the answer trie data structure is not
-shared among threads. The usage of batched scheduling with the FS
-design requires further support since with batched scheduling, answers
-are immediately propagated and we have to ensure that the propagation
-of an answer occurs on all subgoal calls one and only once. To do so,
-we take advantage of the private chaining procedure, presented in the
+inserts it. The usage of batched scheduling with the FS design
+requires further support since with batched scheduling, answers are
+immediately propagated and we have to ensure that the propagation of
+an answer occurs on all subgoal calls one and only once. To do so, we
+take advantage of the private chaining procedure, presented in the
 previous subsection, as a way to keep, for every subgoal call of every
@@
   \STATE $leaf \gets check\_insert\_answer\_trie(ANS, SF)$
-  \IF {$NS\_design\ or\ SS\_design$}
-    \STATE $. . .$                         \COMMENT{without changes}
-  \ELSE                                                  [FS design]
   \STATE $chain \gets check\_insert\_consumer\_chain(leaf, SF)$
@@
   \ENDIF  
-   \ENDIF
 \end{algorithmic}
@@
 evaluation (\emph{ANS}) and the subgoal frame which corresponds to the
-call at hand (\emph{SF}). The \emph{NS\_design}, \emph{SS\_design} and
-\emph{FS\_design} macros define which table design is enabled.
+call at hand (\emph{SF}). The algorithm begins by checking/inserting
+the given \emph{ANS} into the answer trie structure, which will return
+the leaf node for the path representing \emph{ANS} (line 1).
 
-The algorithm begins by checking/inserting the given \emph{ANS} into
-the answer trie structure, which will return the leaf node for the
-path representing \emph{ANS} (line 1). In line 2, it then tests
-whether one of the NS or SS designs are active, and in such a case,
-the algorithm is remains unchanged.
-
-Otherwise, for the FS design (lines 4 to 13), it checks/inserts
-the given \emph{leaf} node into the private consumer chain for the
-current thread, which will return the corresponding chain node. In
-line 6, it then tests whether the chain node already existed in the
-consumer chain, i.e., if it was inserted or not by the current
-check/insert operation in order to return failure (line 7), or it
-proceed with marking the answer \emph{ANS} has found (line 9). At the
-end (lines 10 to 13), it returns failure if local scheduling is active
-(line 11), otherwise, the batched scheduling is active, thus it
-propagates the answer \emph{ANS} (line 13).
+Then, it checks/inserts the given \emph{leaf} node into the private
+consumer chain for the current thread, which will return the
+corresponding chain node (line 2). Next in line 3, it tests whether
+the chain node already existed in the consumer chain, i.e., if it was
+inserted or not by the current check/insert operation in order to
+return failure (line 4), or it proceeds with marking the answer
+\emph{ANS} has found (line 6). At the end (lines 7 to 10), it returns
+failure if local scheduling is active (line 8), otherwise, the batched
+scheduling is active, thus it propagates the answer \emph{ANS} (line
+10).
 

### 2015-04-15 paper.tex
our approach -> begin

@@
 evaluation, a thread views its tables as private but, at the engine
-level, Yap has three designs: (i) The {\bf No-Sharing (NS)} design,
-where each thread allocates fully private tables for each new tabled
-subgoal called during its computation. (ii) The {\bf Subgoal-Sharing
-  (SS)}, where threads share part of the table space and the (iii) The
-{\bf Full-Sharing (FS)}, where threads share the complete table space.
+level, Yap has three designs that vary from a \emph{No-Sharing} (NS)
+design, where each thread allocates fully private tables for each new
+tabled subgoal called during its computation, to a \emph{Full-Sharing}
+(FS), where threads share the complete table space.
 
@@
 with batched scheduling. Experimental results show that despite the
-extra data structures required to extend the FS design with batched
-scheduling, the execution time of the FS design with batched
-scheduling is still quite competitive when comparing against local
-scheduling.
+extra data structures required to combine FS design with batched
+scheduling, the execution time of combination is still quite
+competitive when comparing against the FS with local scheduling.
 
@@
 sections.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{No-Sharing, Subgoal-Sharing and Full-Sharing Designs}
 
@@
 older generator node.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Yap's No-Sharing and Full-Sharing Designs}
+
+\begin{figure}[!ht]
+\centering
+\includegraphics[width=10.5cm]{figures/arch.pdf}
+\caption{The \emph{FS} design with the PAC procedure - (a) private
+  chaining and (b) public chaining}
+\label{fig_tabtries_pcc}
+\end{figure}
 
@@
 compared with the NS design with 1 thread (running with local
-scheduling), for the NS, SS and FS designs (all running TabMalloc),
-when running 1, 8, 16, 24 and 32 threads with local and batched
-scheduling on the five sets of benchmarks. For each design, the table
-has then two columns, a column with {\bf Local} that shows results for
-the local scheduling and a column with {\bf Batched} with results for
+scheduling), for the NS and FS designs (all running TabMalloc), when
+running 1, 8, 16, 24 and 32 threads with local and batched scheduling
+on the five sets of benchmarks. For each design, the table has then
+two columns, a column with {\bf Local} that shows results for the
+local scheduling and a column with {\bf Batched} with results for
 batched scheduling. The overhead results presented in both {\bf Local}
@@
 
+\setlength{\tabcolsep}{12pt}
+
 \begin{table}[!ht]
@@
 \caption{Overhead ratios, when compared with the NS design with $1$
-  thread (running with local scheduling without TabMalloc) for the NS,
-  SS, FS designs (with TabMalloc), when running 1, 8, 16, 24 and 32
+  thread (running with local scheduling without TabMalloc) for the NS
+  and FS designs (with TabMalloc), when running 1, 8, 16, 24 and 32
   threads with local and batched scheduling on the five sets of
@@
 %\scalebox{1.00} {
-\begin{tabular}{ll|cc|cc|cc}
+\begin{tabular}{ll|cc|cc}
 \hline\hline
@@
 \multicolumn{2}{c|}{\multirow{1}{*}{\bf NS}} &
-\multicolumn{2}{c}{\multirow{1}{*}{\bf SS}} & 
-\multicolumn{2}{|c}{\multirow{1}{*}{\bf FS}}\\ \cline{3-8}
+\multicolumn{2}{|c}{\multirow{1}{*}{\bf FS}}\\ \cline{3-6}
 & 
 & \multicolumn{1}{c}{\bf Local}
-& \multicolumn{1}{c}{\bf Batched}
-& \multicolumn{1}{|c}{\bf Local}
 & \multicolumn{1}{c}{\bf Batched}
@@
 \multirow{4}{*}{\bf 1}
-& {\bf Min }& {\bf 0.53}& 0.55& {\bf 0.54}& 0.55& 1.01& {\bf 0.95}\\
-& {\bf Avg }& {\bf 0.78}& 0.82& {\bf 0.84}& 0.90& {\bf 1.30}& 1.46\\
-& {\bf Max }& 1.06& {\bf 1.05}& {\bf 1.04}& {\bf 1.04}& {\bf 1.76}& 2.33\\
-& {\bf StD }& 0.15& 0.14& 0.17& 0.16& 0.22& 0.44\\
+& {\bf Min }& {\bf 0.53}& 0.55& 1.01& {\bf 0.95}\\
+& {\bf Avg }& {\bf 0.78}& 0.82& {\bf 1.30}& 1.46\\
+& {\bf Max }& 1.06& {\bf 1.05}& {\bf 1.76}& 2.33\\
+& {\bf StD }& 0.15& 0.14& 0.22& 0.44\\
 \hline
 \multirow{4}{*}{\bf 8}
-& {\bf Min }& 0.66& {\bf 0.63}& 0.66& {\bf 0.63}& 1.16&{\bf  0.99}\\
-& {\bf Avg }& {\bf 0.85}& 0.88& {\bf 0.92}& 0.93& {\bf 1.88}& 1.95\\
-& {\bf Max }& {\bf 1.12}& 1.14& 1.20& {\bf 1.15}& {\bf 2.82}& 3.49\\
-& {\bf StD }& 0.13& 0.14& 0.15& 0.14& 0.60& 0.79\\
+& {\bf Min }& 0.66& {\bf 0.63}& 1.16&{\bf  0.99}\\
+& {\bf Avg }& {\bf 0.85}& 0.88& {\bf 1.88}& 1.95\\
+& {\bf Max }& {\bf 1.12}& 1.14& {\bf 2.82}& 3.49\\
+& {\bf StD }& 0.13& 0.14& 0.60& 0.79\\
 \hline
 \multirow{4}{*}{\bf 16}
-& {\bf Min }& 0.85& {\bf 0.75}& 0.82& {\bf 0.77}& 1.17& {\bf 1.06}\\
-& {\bf Avg }& {\bf 0.98}& 1.00& {\bf 1.04}& 1.05& {\bf 1.97}& 2.08\\
-& {\bf Max }& {\bf 1.16}& 1.31& 1.31& {\bf 1.28}& {\bf 3.14}& 3.69\\
-& {\bf StD }& 0.09& 0.17& 0.12& 0.13& 0.65& 0.83\\
+& {\bf Min }& 0.85& {\bf 0.75}& 1.17& {\bf 1.06}\\
+& {\bf Avg }& {\bf 0.98}& 1.00& {\bf 1.97}& 2.08\\
+& {\bf Max }& {\bf 1.16}& 1.31& {\bf 3.14}& 3.69\\
+& {\bf StD }& 0.09& 0.17& 0.65& 0.83\\
 \hline
 \multirow{4}{*}{\bf 24}
-& {\bf Min }& {\bf 0.91}& 0.93& 1.02& {\bf 0.98}& 1.16& {\bf 1.09}\\
-& {\bf Avg }& {\bf 1.15}& 1.16& 1.22& {\bf 1.19}& {\bf 2.06}& 2.19\\
-& {\bf Max }& 1.72& {\bf 1.60}& 1.81& {\bf 1.61}& {\bf 3.49}& 4.08\\
-& {\bf StD }& 0.20& 0.21& 0.18& 0.16& 0.70& 0.91\\
+& {\bf Min }& {\bf 0.91}& 0.93& 1.16& {\bf 1.09}\\
+& {\bf Avg }& {\bf 1.15}& 1.16& {\bf 2.06}& 2.19\\
+& {\bf Max }& 1.72& {\bf 1.60}& {\bf 3.49}& 4.08\\
+& {\bf StD }& 0.20& 0.21& 0.70& 0.91\\
 \hline
 \multirow{4}{*}{\bf 32}
-& {\bf Min }& 1.05& {\bf 1.04}& {\bf 1.07}& 1.12& 1.33& {\bf 1.26}\\
-& {\bf Avg }& 1.51& {\bf 1.49}& 1.54& {\bf 1.51}& {\bf 2.24}& 2.41\\
-& {\bf Max }& {\bf 2.52}& 2.63& {\bf 2.52}& 2.62& {\bf 3.71}& 4.51\\
-& {\bf StD }& 0.45& 0.45& 0.42& 0.43& 0.74& 1.02\\
+& {\bf Min }& 1.05& {\bf 1.04}& 1.33& {\bf 1.26}\\
+& {\bf Avg }& 1.51& {\bf 1.49}& {\bf 2.24}& 2.41\\
+& {\bf Max }& {\bf 2.52}& 2.63& {\bf 3.71}& 4.51\\
+& {\bf StD }& 0.45& 0.45& 0.74& 1.02\\
 \hline\hline
@@
 one thread, on average, local scheduling is sightly better than
-batched on the three designs. For the NS design we have $0.78$ and
-$0.82$, for the SS design we have $0.84$ and $0.90$ and for the FS
-design we have $1.30$ and $1.46$ average overhead ratios, for the
-local and batched scheduling strategies, respectively. As we scale the
-number of threads, one can observe that, for the NS and SS designs
-both scheduling strategies have similar minimum, average and maximum
-overhead ratios. For the FS design, the best minimum overhead ratio is
-always for batched scheduling. For the average and maximum overhead
-ratio, local scheduling is always better than batched scheduling. In
-summary, we can say that both the local and batched scheduling
-strategies have similar overhead results on worst case scenarios for
-the NS, SS and FS designs.
+batched on the two designs. For the NS design we have $0.78$ and
+$0.82$ and for the FS design we have $1.30$ and $1.46$ average
+overhead ratios, for the local and batched scheduling strategies,
+respectively. As we scale the number of threads, one can observe that,
+for the NS design both scheduling strategies have similar minimum,
+average and maximum overhead ratios. For the FS design, the best
+minimum overhead ratio on $1$, $8$, $16$, $24$ and $32$ threads is
+always the batched scheduling. For the average and maximum overhead
+ratio, local scheduling is always better than batched
+scheduling. However, for the average ratio, the results between both
+strategies is quite close. We consider this to be quite significant,
+once the FS design with batched scheduling requires the support of the
+PAC procedure. For the maximum overhead ratio, the difference between
+both strategies is higher than for the average overhead ratio, however
+we hope to reduce the difference with further improvements in the PAC
+procedure.
 

### 2015-04-15 paper.tex
batched scheduling ->ok for now

@@
 were the only Prolog systems that were able to combine both
-multithreading with tabling. Yap's threads have their own execution
-stacks and only share the code area where predicates, records, flags
-and other global non-backtrackable data are stored. For tabled
-evaluation, a thread views its tables as private but, at the engine
-level, Yap has three designs that vary from a \emph{No-Sharing} (NS)
-design, where each thread allocates fully private tables for each new
-tabled subgoal called during its computation, to a \emph{Full-Sharing}
-(FS), where threads share the complete table space.
+multithreading with tabling. Yap implements a SWI-Prolog compatible
+multi-threading library~\cite{Wielemaker-03}. Yap's threads have their
+own execution stacks and only share the code area where predicates,
+records, flags and other global non-backtrackable data are stored. For
+tabled evaluation, a thread views its tables as private but, at the
+engine level, Yap has three designs that vary from a \emph{No-Sharing}
+(NS) design, where each thread allocates fully private tables for each
+new tabled subgoal called during its computation, to a
+\emph{Full-Sharing} (FS), where threads share the complete table
+space.
 
@@
 sections.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Table Space Organization}
+
+The basic idea behind tabling is straightforward: programs are
+evaluated by storing answers for tabled subgoals in an appropriate
+data space, called the \emph{table space}. Similar calls to tabled
+subgoals are not re-evaluated against the program clauses, instead
+they are resolved by consuming the answers already stored in their
+table entries. During this process, as further new answers are found,
+they are stored in their tables and later returned to all similar
+calls.
+
+\begin{wrapfigure}{r}{4.5cm}
+\vspace{-\intextsep}
+\centering
+\includegraphics[width=4.5cm]{figures/table-space.pdf}
+\caption{Table space organization}
+\label{fig_table_space}
+\vspace{-0.5\intextsep}
+\end{wrapfigure}
+
+Figure~\ref{fig_table_space} shows the table space organization of the
+Yap system. At the entry point we have the \emph{table entry} data
+structure. This structure is allocated when a tabled predicate is
+being compiled, so that a pointer to the table entry can be included
+in its compiled code. This guarantees that further calls to the
+predicate will access the table space starting from the same
+point. Below the table entry, we have the \emph{subgoal trie
+  structure}. Each different tabled subgoal call to the predicate at
+hand corresponds to a unique path through the subgoal trie structure,
+always starting from the table entry, passing by several subgoal trie
+data units, the \emph{subgoal trie nodes}, and reaching a leaf data
+structure, the \emph{subgoal frame}. The subgoal frame stores
+additional information about the subgoal and acts like an entry point
+to the \emph{answer trie structure}. Each unique path through the
+answer trie data units, the \emph{answer trie nodes}, corresponds to a
+different answer to the entry subgoal.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Yap's Multithreaded Tabling Support}
+
+Yap implements a SWI-Prolog compatible multi-threading
+library~\cite{Wielemaker-03}. Like in SWI-Prolog, Yap's threads have
+their own execution stacks and only share the code area where
+predicates, records, flags and other global non-backtrackable data are
+stored.
+
+In a multithreaded tabling evaluation, a thread views its tables as
+private but, at the engine level, Yap has three designs that vary from
+a \emph{No-Sharing} (NS) design, where each thread allocates fully
+private tables for each new tabled subgoal called during its
+computation, to a \emph{Full-Sharing} (FS), where threads share the
+complete table space. Figure~\ref{fig_table_space} shows Yap's
+multithreaded tabling support with an abstraction layer to which
+threads interact (the threads represented in the bottom of the
+figure), and at the engine level the configuration of the NS and FS
+designs in a situation where several different threads evaluated the
+same tabled subgoal call $i$.
+
+In the NS design, each thread allocates fully private tables for each
+new subgoal called during its computation. One can observe that the
+table entry data structure still stores the common information for the
+predicate (such as the arity or the evaluation strategy), and then
+each thread $t$ has its own cell $T_t$ inside a \emph{bucket array}
+which points to the private data structures.
+
+\begin{figure}[!ht]
+\centering
+\includegraphics[width=10.5cm]{figures/yap-mt.pdf}
+\caption{Yap's multithreaded tabled space using the NS and FS designs}
+\label{fig_yap_mt_support}
+\end{figure}
+
+In the FS design, part of the subgoal frame information (the
+\emph{subgoal entry} data structure) and the answer trie structure are
+now shared among all threads. The previous subgoal frame data
+structure was split into two: the \emph{subgoal entry} stores common
+information for the subgoal call (such as the pointer to the shared
+answer trie structure); the remaining information (the \emph{subgoal
+  frame} data structure) remains private to each thread.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Yap's No-Sharing and Full-Sharing Designs}
+\section{Full-Sharing with Batched Scheduling}
 
-\begin{figure}[!ht]
-\centering
-\includegraphics[width=10.5cm]{figures/arch.pdf}
-\caption{The \emph{FS} design with the PAC procedure - (a) private
-  chaining and (b) public chaining}
-\label{fig_tabtries_pcc}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Full-Sharing with Batched Scheduling}
+In this section, we show how we have extended the FS design to support
+batched scheduling.
 
@@
 supporting the implementation of the PAC procedure during the
-evaluation of a tabled subgoal call $P_{i.j}$ using the FS design. The
-FS design, uses a subgoal entry data structure to store common
+evaluation of a tabled subgoal call $i$ using the FS design. The FS
+design, uses a subgoal entry data structure to store common
 information for a subgoal call and a subgoal frame (\emph{SF}) data
@@
 threads, $T_1$ and $T_{k-1}$, are sharing the same subgoal entry call
-$P_{i.j}$ when the subgoal is still under evaluation, i.e., the
-subgoal is not yet complete. The current state of the evaluation shows
-an answer trie with $3$ answers found for the subgoal call
-$P_{i.j}$. For the sake of simplicity, we are omitting the internal
-answer trie nodes and we are only showing the leaf nodes nodes $LN_1$,
-$LN_2$ and $LN_3$, in the figure. With the PAC procedure, the leaf
-nodes are not chained in the \emph{AT} data structure. Now, the
-chaining process is done privately, and for that, we use the subgoal
-frame structure of each thread. On the subgoal frame structure we
-added a new field, called \emph{consumer}, to store the answers found
-within the execution of the thread. In order to minimize the impact of
-the \emph{PAC} optimization, each node within the new consumer answer
-structure has two fields: (i) an entry pointer, which points to the
-corresponding leaf node in the answer trie data structure; (ii) a next
-pointer to chain the answers within the consumer structure. To
-maintain a good performance, when the number of nodes exceeds a
-certain threshold, we use a hash trie mechanism design similar to the
-one presented in the work~\cite{Areias-ijpp15}. However, since this
-mechanism is private to each thread, it does not require any of the
-tools that were necessary to support concurrency. In particular, on
-each hash trie level, we have removed the tools necessary to support
-concurrency, such as useless pointers and compare-and-swap
-operations. We have chosen this hashing mechanism, because it showed a
-good balance between lookup and insert
-operations~\cite{Areias-ijpp15}, but the major reason was mostly
-because of the integration in the TabMalloc memory
+$i$ when the subgoal is still under evaluation, i.e., the subgoal is
+not yet complete. The current state of the evaluation shows an answer
+trie with $3$ answers found for the subgoal call $i$. For the sake of
+simplicity, we are omitting the internal answer trie nodes and we are
+only showing the leaf nodes nodes $LN_1$, $LN_2$ and $LN_3$, in the
+figure. With the PAC procedure, the leaf nodes are not chained in the
+\emph{AT} data structure. Now, the chaining process is done privately,
+and for that, we use the subgoal frame structure of each thread. On
+the subgoal frame structure we added a new field, called
+\emph{consumer}, to store the answers found within the execution of
+the thread. In order to minimize the impact of the \emph{PAC}
+optimization, each node within the new consumer answer structure has
+two fields: (i) an entry pointer, which points to the corresponding
+leaf node in the answer trie data structure; (ii) a next pointer to
+chain the answers within the consumer structure. To maintain a good
+performance, when the number of nodes exceeds a certain threshold, we
+use a hash trie mechanism design similar to the one presented in the
+work~\cite{Areias-ijpp15}. However, since this mechanism is private to
+each thread, it does not require any of the tools that were necessary
+to support concurrency. In particular, on each hash trie level, we
+have removed the tools necessary to support concurrency, such as
+useless pointers and compare-and-swap operations. We have chosen this
+hashing mechanism, because it showed a good balance between lookup and
+insert operations~\cite{Areias-ijpp15}, but the major reason was
+mostly because of the integration in the TabMalloc memory
 allocator~\cite{Areias-12b}.
@@
 
-We used five sets of benchmarks. The \emph{Large Joins} and
-\emph{WordNet} sets were obtained from the OpenRuleBench
-project~\cite{Liang-09}; the \emph{Model Checking} set includes three
-different specifications and transition relation graphs usually used
-in model checking applications; the \emph{Path Left} and \emph{Path
-  Right} sets implement two recursive definitions of the well-known
-$path/2$ predicate, that computes the transitive closure in a graph,
-using several different configurations of $edge/2$ facts. We
-experimented graphs with the \emph{BTree} configuration with depth 17,
-the \emph{Pyramid} and \emph{Cycle} configurations with depth 2000 and
-the \emph{Grid} configuration with depth 35.
+%% We used five sets of benchmarks. The \emph{Large Joins} and
+%% \emph{WordNet} sets were obtained from the OpenRuleBench
+%% project~\cite{Liang-09}; the \emph{Model Checking} set includes three
+%% different specifications and transition relation graphs usually used
+%% in model checking applications; the \emph{Path Left} and \emph{Path
+%%   Right} sets implement two recursive definitions of the well-known
+%% $path/2$ predicate, that computes the transitive closure in a graph,
+%% using several different configurations of $edge/2$ facts. We
+%% experimented graphs with the \emph{BTree} configuration with depth 17,
+%% the \emph{Pyramid} and \emph{Cycle} configurations with depth 2000 and
+%% the \emph{Grid} configuration with depth 35.
 
-In order to have a deeper insight on the behavior of each benchmark,
-and therefore clarify some of the results that are presented next, we
-first characterize the benchmarks. The columns in
-Table~\ref{tab_benchs} have the following meaning: (i) {\bf calls:} is
-the number of different calls to tabled subgoals. It corresponds to
-the number of paths in the subgoal tries. (ii) {\bf trie nodes:} is
-the total number of trie nodes allocated in the corresponding
-subgoal/answer trie structures. (iii) {\bf trie depth:} is the
-minimum/average/maximum number of trie node levels required to
-represent a path in the corresponding subgoal/answer trie
-structures. Trie structures with smaller average values are more
-amenable to higher lock contention. (iv) {\bf unique:} is the number
-of different tabled answers found. It corresponds to the number of
-paths in the answer tries. (v) {\bf repeated:} is the number of
-redundant tabled answers found. (vi) {\bf NS:} is the average
-execution time, in seconds, of ten runs for 1 thread with the NS
-design without the TabMalloc memory allocator. In what follows, we
-will use these times as the base times when computing the overhead
-ratios for the other designs.
+%% In order to have a deeper insight on the behavior of each benchmark,
+%% and therefore clarify some of the results that are presented next, we
+%% first characterize the benchmarks. The columns in
+%% Table~\ref{tab_benchs} have the following meaning: (i) {\bf calls:} is
+%% the number of different calls to tabled subgoals. It corresponds to
+%% the number of paths in the subgoal tries. (ii) {\bf trie nodes:} is
+%% the total number of trie nodes allocated in the corresponding
+%% subgoal/answer trie structures. (iii) {\bf trie depth:} is the
+%% minimum/average/maximum number of trie node levels required to
+%% represent a path in the corresponding subgoal/answer trie
+%% structures. Trie structures with smaller average values are more
+%% amenable to higher lock contention. (iv) {\bf unique:} is the number
+%% of different tabled answers found. It corresponds to the number of
+%% paths in the answer tries. (v) {\bf repeated:} is the number of
+%% redundant tabled answers found. (vi) {\bf NS:} is the average
+%% execution time, in seconds, of ten runs for 1 thread with the NS
+%% design without the TabMalloc memory allocator. In what follows, we
+%% will use these times as the base times when computing the overhead
+%% ratios for the other designs.
 
-By observing Table~\ref{tab_benchs}, the \emph{Mondial} benchmark,
-from the \emph{Large Joins} set, and the three \emph{Model
-  Checking} benchmarks seem to be the benchmarks least amenable to
-lock contention since they are the ones that find less unique answers
-and that have the deepest trie structures. In this regard, the
-\emph{Path Left} and \emph{Path Right} sets correspond to the
-opposite case. They find a huge number of answers and have very
-shallow trie structures. On the other hand, the \emph{WordNet} and
-\emph{Path Right} sets have the benchmarks with the largest number
-of different subgoal calls, which can reduce the probability of lock
-contention because answers can be found for different subgoal calls
-and therefore be inserted with minimum overlap. On the opposite side
-are the \emph{Join2} benchmark, from the \emph{Large Joins} set,
-and the \emph{Path Left} benchmarks, which have only a single tabled
-subgoal call.
+%% By observing Table~\ref{tab_benchs}, the \emph{Mondial} benchmark,
+%% from the \emph{Large Joins} set, and the three \emph{Model
+%%   Checking} benchmarks seem to be the benchmarks least amenable to
+%% lock contention since they are the ones that find less unique answers
+%% and that have the deepest trie structures. In this regard, the
+%% \emph{Path Left} and \emph{Path Right} sets correspond to the
+%% opposite case. They find a huge number of answers and have very
+%% shallow trie structures. On the other hand, the \emph{WordNet} and
+%% \emph{Path Right} sets have the benchmarks with the largest number
+%% of different subgoal calls, which can reduce the probability of lock
+%% contention because answers can be found for different subgoal calls
+%% and therefore be inserted with minimum overlap. On the opposite side
+%% are the \emph{Join2} benchmark, from the \emph{Large Joins} set,
+%% and the \emph{Path Left} benchmarks, which have only a single tabled
+%% subgoal call.
 
-\begin{table}[ht]
-\centering
-\caption{Characteristics of the benchmark programs}
-\scalebox{0.82} { 
-\begin{tabular}{lrrccrrrcc}
-\hline\hline
-\multirow{2}{*}{\bf Bench}
-& \multicolumn{3}{c}{\bf Tabled Subgoals}
-&
-& \multicolumn{4}{c}{\bf Tabled Answers}
-& \multicolumn{1}{c}{\bf Time (sec)} \\ \cline{2-4}\cline{6-9}
-& \multicolumn{1}{c}{\bf calls}
-& \multicolumn{1}{c}{\bf trie nodes}
-& \multicolumn{1}{c}{\bf trie depth}
-&
-& \multicolumn{1}{c}{\bf unique}
-& \multicolumn{1}{c}{\bf repeated}
-& \multicolumn{1}{c}{\bf trie nodes}
-& \multicolumn{1}{c}{\bf trie depth}
-& \multicolumn{1}{c}{\bf NS} \\
-\hline
-\multicolumn{10}{l}{\bf Large Joins} \\
-\bf{Join2}    &       1 &       6 & 5/5/5 && 2,476,099 &         0 &
-2,613,660 &    5/5/5 &  2.85 \\
-\bf{Mondial}  &      35 &      42 & 3/4/4 &&     2,664 & 2,452,890 &
-   14,334 &    6/7/7 &  0.84 \\
-\hline
-\multicolumn{10}{l}{\bf WordNet} \\
-\bf{Clusters} & 117,659 & 235,319 & 2/2/2 &&   166,877 &   161,853 &
-284,536 &    1/1/1 &   0.83 \\
-\bf{Holo}     & 117,657 & 235,315 & 2/2/2 &&    74,838 &        54 &
-192,495 &    1/1/1 &   0.75 \\
-\bf{Hyper}    & 117,657 & 235,315 & 2/2/2 &&   698,472 &     8,658 &
-816,129 &    1/1/1 &   1.42 \\
-\bf{Hypo}     & 117,657 & 117,659 & 2/2/2 &&   698,472 &    20,341 &
-816,129 &    1/1/1 &   1.53 \\
-\bf{Mero}     & 117,657 & 117,659 & 2/2/2 &&    74,838 &        13 &
-192,495 &    1/1/1 &   0.74 \\
-\bf{Tropo}    & 117,657 & 235,315 & 2/2/2 &&       472 &         0 &
-118,129 &    1/1/1 &   0.66 \\
-\hline
-\multicolumn{10}{l}{\bf Model Checking} \\
-\bf{IProto}   &       1 &       6 & 5/5/5 &&   134,361 &   385,423 &
-1,554,896 &  4/51/67 & 2.70 \\
-\bf{Leader}   &       1 &       5 & 4/4/4 &&     1,728 &   574,786 &
-   41,788 & 15/80/97 & 3.51  \\
-\bf{Sieve}    &       1 &       7 & 6/6/6 &&       380 & 1,386,181 &
-    8,624 & 21/53/58 & 18.50 \\
-\hline
-\multicolumn{10}{l}{\bf Path Left} \\
-\bf{BTree}    &       1 &       3 & 2/2/2 && 1,966,082 &         0 &
-2,031,618 &    2/2/2 &   1.53 \\
-\bf{Cycle}    &       1 &       3 & 2/2/2 && 4,000,000 &     2,000 &
-4,002,001 &    2/2/2 &   3.52 \\
-\bf{Grid}     &       1 &       3 & 2/2/2 && 1,500,625 & 4,335,135 &
-1,501,851 &    2/2/2 &  1.93 \\
-\bf{Pyramid}  &       1 &       3 & 2/2/2 && 3,374,250 & 1,124,250 &
-3,377,250 &    2/2/2 &   3.08 \\
-\hline
-\multicolumn{10}{l}{\bf Path Right} \\
-\bf{BTree}    & 131,071 & 262,143 & 2/2/2 && 3,801,094 &         0 &
-3,997,700 &    1/2/2 &  2.33 \\
-\bf{Cycle}    &   2,001 &   4,003 & 2/2/2 && 8,000,000 &     4,000 &
-8,004,001 &    1/2/2 &  3.55 \\
-\bf{Grid}     &   1,226 &   2,453 & 2/2/2 && 3,001,250 & 8,670,270 &
-3,003,701 &    1/2/2 &  2.32 \\
-\bf{Pyramid}  &   3,000 &   6,001 & 2/2/2 && 6,745,501 & 2,247,001 &
-6,751,500 &    1/2/2 &  3.17 \\
-\hline\hline
-\end{tabular}}
-\label{tab_benchs}
-\end{table}
+%% \begin{table}[ht]
+%% \centering
+%% \caption{Characteristics of the benchmark programs}
+%% \scalebox{0.82} { 
+%% \begin{tabular}{lrrccrrrcc}
+%% \hline\hline
+%% \multirow{2}{*}{\bf Bench}
+%% & \multicolumn{3}{c}{\bf Tabled Subgoals}
+%% &
+%% & \multicolumn{4}{c}{\bf Tabled Answers}
+%% & \multicolumn{1}{c}{\bf Time (sec)} \\ \cline{2-4}\cline{6-9}
+%% & \multicolumn{1}{c}{\bf calls}
+%% & \multicolumn{1}{c}{\bf trie nodes}
+%% & \multicolumn{1}{c}{\bf trie depth}
+%% &
+%% & \multicolumn{1}{c}{\bf unique}
+%% & \multicolumn{1}{c}{\bf repeated}
+%% & \multicolumn{1}{c}{\bf trie nodes}
+%% & \multicolumn{1}{c}{\bf trie depth}
+%% & \multicolumn{1}{c}{\bf NS} \\
+%% \hline
+%% \multicolumn{10}{l}{\bf Large Joins} \\
+%% \bf{Join2}    &       1 &       6 & 5/5/5 && 2,476,099 &         0 &
+%% 2,613,660 &    5/5/5 &  2.85 \\
+%% \bf{Mondial}  &      35 &      42 & 3/4/4 &&     2,664 & 2,452,890 &
+%%    14,334 &    6/7/7 &  0.84 \\
+%% \hline
+%% \multicolumn{10}{l}{\bf WordNet} \\
+%% \bf{Clusters} & 117,659 & 235,319 & 2/2/2 &&   166,877 &   161,853 &
+%% 284,536 &    1/1/1 &   0.83 \\
+%% \bf{Holo}     & 117,657 & 235,315 & 2/2/2 &&    74,838 &        54 &
+%% 192,495 &    1/1/1 &   0.75 \\
+%% \bf{Hyper}    & 117,657 & 235,315 & 2/2/2 &&   698,472 &     8,658 &
+%% 816,129 &    1/1/1 &   1.42 \\
+%% \bf{Hypo}     & 117,657 & 117,659 & 2/2/2 &&   698,472 &    20,341 &
+%% 816,129 &    1/1/1 &   1.53 \\
+%% \bf{Mero}     & 117,657 & 117,659 & 2/2/2 &&    74,838 &        13 &
+%% 192,495 &    1/1/1 &   0.74 \\
+%% \bf{Tropo}    & 117,657 & 235,315 & 2/2/2 &&       472 &         0 &
+%% 118,129 &    1/1/1 &   0.66 \\
+%% \hline
+%% \multicolumn{10}{l}{\bf Model Checking} \\
+%% \bf{IProto}   &       1 &       6 & 5/5/5 &&   134,361 &   385,423 &
+%% 1,554,896 &  4/51/67 & 2.70 \\
+%% \bf{Leader}   &       1 &       5 & 4/4/4 &&     1,728 &   574,786 &
+%%    41,788 & 15/80/97 & 3.51  \\
+%% \bf{Sieve}    &       1 &       7 & 6/6/6 &&       380 & 1,386,181 &
+%%     8,624 & 21/53/58 & 18.50 \\
+%% \hline
+%% \multicolumn{10}{l}{\bf Path Left} \\
+%% \bf{BTree}    &       1 &       3 & 2/2/2 && 1,966,082 &         0 &
+%% 2,031,618 &    2/2/2 &   1.53 \\
+%% \bf{Cycle}    &       1 &       3 & 2/2/2 && 4,000,000 &     2,000 &
+%% 4,002,001 &    2/2/2 &   3.52 \\
+%% \bf{Grid}     &       1 &       3 & 2/2/2 && 1,500,625 & 4,335,135 &
+%% 1,501,851 &    2/2/2 &  1.93 \\
+%% \bf{Pyramid}  &       1 &       3 & 2/2/2 && 3,374,250 & 1,124,250 &
+%% 3,377,250 &    2/2/2 &   3.08 \\
+%% \hline
+%% \multicolumn{10}{l}{\bf Path Right} \\
+%% \bf{BTree}    & 131,071 & 262,143 & 2/2/2 && 3,801,094 &         0 &
+%% 3,997,700 &    1/2/2 &  2.33 \\
+%% \bf{Cycle}    &   2,001 &   4,003 & 2/2/2 && 8,000,000 &     4,000 &
+%% 8,004,001 &    1/2/2 &  3.55 \\
+%% \bf{Grid}     &   1,226 &   2,453 & 2/2/2 && 3,001,250 & 8,670,270 &
+%% 3,003,701 &    1/2/2 &  2.32 \\
+%% \bf{Pyramid}  &   3,000 &   6,001 & 2/2/2 && 6,745,501 & 2,247,001 &
+%% 6,751,500 &    1/2/2 &  3.17 \\
+%% \hline\hline
+%% \end{tabular}}
+%% \label{tab_benchs}
+%% \end{table}
 
-Note that for the moment our goal is to evaluate the robustness of our
-implementation when exposed to \emph{worst case} scenarios. By
-focusing on \emph{worst case} scenarios, we show the lowest bounds in
-terms of performance that each design might achieve when applied/used
-with other real world applications/programs. Moreover, by testing the
-framework with worst case scenarios, we avoid the peculiarities of the
-program at hand and we try to focus on measuring the real value of our
-designs. Thus, we will follow a common approach to create worst case
-scenarios and we will run all threads starting with the same query
-goal. By doing this, it is expected that all threads will access the
-table space, to check/insert for subgoals and answers, at similar
-times, thus causing a huge stress on the same critical regions. To put
-the results in perspective, we experimented with $1$, $8$, $16$, $24$
-and $32$ threads (the maximum number of cores available in our
-machine) with local and batched scheduling on the five sets of
-benchmarks presented in the early subsection.
+%% Note that for the moment our goal is to evaluate the robustness of our
+%% implementation when exposed to \emph{worst case} scenarios. By
+%% focusing on \emph{worst case} scenarios, we show the lowest bounds in
+%% terms of performance that each design might achieve when applied/used
+%% with other real world applications/programs. Moreover, by testing the
+%% framework with worst case scenarios, we avoid the peculiarities of the
+%% program at hand and we try to focus on measuring the real value of our
+%% designs. Thus, we will follow a common approach to create worst case
+%% scenarios and we will run all threads starting with the same query
+%% goal. By doing this, it is expected that all threads will access the
+%% table space, to check/insert for subgoals and answers, at similar
+%% times, thus causing a huge stress on the same critical regions. To put
+%% the results in perspective, we experimented with $1$, $8$, $16$, $24$
+%% and $32$ threads (the maximum number of cores available in our
+%% machine) with local and batched scheduling on the five sets of
+%% benchmarks presented in the early subsection.
 

### 2015-04-15 paper.tex
batched scheduling ->ok for now

@@
 
-In this paper we discuss a novel approach for solving the problem of
-supporting multithreaded batched scheduling in the FS design and we
-present a performance analysis comparison between local scheduling
-with batched scheduling. Experimental results show that despite the
-extra data structures required to combine FS design with batched
-scheduling, the execution time of combination is still quite
-competitive when comparing against the FS with local scheduling.
+In this paper we discuss a novel approach for combining the FS design
+with batched scheduling and we present a performance analysis
+comparison between local scheduling with batched
+scheduling. Experimental results show that despite the extra data
+structures required to combine FS design with batched scheduling, the
+execution time of combination is still quite competitive when
+comparing against the FS with local scheduling.
 
@@
 \centering
-\includegraphics[width=10.5cm]{figures/yap-mt.pdf}
+\includegraphics[width=9.5cm]{figures/yap-mt.pdf}
 \caption{Yap's multithreaded tabled space using the NS and FS designs}
@@
 In this section, we show how we have extended the FS design to support
-batched scheduling.
+batched scheduling. The usage of batched scheduling with the FS design
+requires further support since with batched scheduling, answers are
+immediately propagated and we have to ensure that the propagation of
+an answer occurs on all subgoal calls one and only once. To do so,
+next we show a procedure that, for every subgoal call of every thread,
+keeps track of all the answers that were already propagated.
 
@@
 figure. With the PAC procedure, the leaf nodes are not chained in the
-\emph{AT} data structure. Now, the chaining process is done privately,
-and for that, we use the subgoal frame structure of each thread. On
-the subgoal frame structure we added a new field, called
+answer trie data structure. Now, the chaining process is done
+privately, and for that, we use the subgoal frame structure of each
+thread. On the subgoal frame structure we added a new field, called
 \emph{consumer}, to store the answers found within the execution of
@@
 already in the corresponding answer trie structure and, if not,
-inserts it. The usage of batched scheduling with the FS design
-requires further support since with batched scheduling, answers are
-immediately propagated and we have to ensure that the propagation of
-an answer occurs on all subgoal calls one and only once. To do so, we
-take advantage of the private chaining procedure, presented in the
-previous subsection, as a way to keep, for every subgoal call of every
-thread, track of all the answers that were already propagated. This
-requires minor changes to the \emph{tabled new answer} tabling
+inserts it. This requires minor changes to the \emph{tabled new
+  answer} tabling
 operation. Algorithm~\ref{alg_table_new_answer_batched} shows how we

### 2015-04-15 paper.tex
batched scheduling ->ok for now

@@
 \begin{figure}[!ht]
+\vspace{-\intextsep}
 \centering
@@
 \label{fig_yap_mt_support}
+\vspace{-1.0\intextsep}
 \end{figure}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Batched Scheduling}
+\subsection{Scheduling Strategies}
 
@@
 to return answers only after the completion point has been
-reached~\cite{Freire-96}. In other words, the local scheduling tries
-to keep a CDS as minimal as possible. When new answers are found, they
-are added to the table space and the computation fails as consequence,
-tabled subgoals inside a CDS propagate their answers to outside the
-CDS only after its completion point is found. Local scheduling causes
-a sooner completion of subgoals, which creates less complex
-dependencies between them.
+reached~\cite{Freire-96}. %% In other words, the local scheduling tries
+%% to keep a CDS as minimal as possible. When new answers are found, they
+%% are added to the table space and the computation fails as consequence,
+%% tabled subgoals inside a CDS propagate their answers to outside the
+%% CDS only after its completion point is found.
+Local scheduling causes a sooner completion of subgoals, which creates
+less complex dependencies between them.
 
@@
 \label{fig_tabtries_pcc}
+\vspace{-1.0\intextsep}
 \end{figure}
@@
 
-\begin{algorithm} [ht]
+\begin{algorithm} [!ht]
 \caption{tabled\_new\_answer(answer ANS, subgoal frame SF)}
@@
 
-%% We used five sets of benchmarks. The \emph{Large Joins} and
-%% \emph{WordNet} sets were obtained from the OpenRuleBench
-%% project~\cite{Liang-09}; the \emph{Model Checking} set includes three
-%% different specifications and transition relation graphs usually used
-%% in model checking applications; the \emph{Path Left} and \emph{Path
-%%   Right} sets implement two recursive definitions of the well-known
-%% $path/2$ predicate, that computes the transitive closure in a graph,
-%% using several different configurations of $edge/2$ facts. We
-%% experimented graphs with the \emph{BTree} configuration with depth 17,
-%% the \emph{Pyramid} and \emph{Cycle} configurations with depth 2000 and
-%% the \emph{Grid} configuration with depth 35.
+We used five sets of benchmarks. The \emph{Large Joins} and
+\emph{WordNet} sets were obtained from the OpenRuleBench
+project~\cite{Liang-09}; the \emph{Model Checking} set includes three
+different specifications and transition relation graphs usually used
+in model checking applications; the \emph{Path Left} and \emph{Path
+  Right} sets implement two recursive definitions of the well-known
+$path/2$ predicate, that computes the transitive closure in a graph,
+using several different configurations of $edge/2$ facts. We
+experimented graphs with the \emph{BTree} configuration with depth 17,
+the \emph{Pyramid} and \emph{Cycle} configurations with depth 2000 and
+the \emph{Grid} configuration with depth 35.
 
@@
 \begin{table}[!ht]
+\vspace{-\intextsep}
 \centering
@@
   Average and Maximum are in bold)}
+\vspace{-0.5\intextsep}
 %\scalebox{1.00} {
@@
 \label{tab_batched_overhead}
+\vspace{-0.7\intextsep}
 \end{table}
@@
 \section{Conclusions and Further Work}
+.........
 

### 2015-04-16 paper.tex
removing stuff

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Full-Sharing with Batched Scheduling}
+\section{Extending Full-Sharing with Batched Scheduling}
 
 In this section, we show how we have extended the FS design to support
-batched scheduling. The usage of batched scheduling with the FS design
-requires further support since with batched scheduling, answers are
-immediately propagated and we have to ensure that the propagation of
-an answer occurs on all subgoal calls one and only once. To do so,
-next we show a procedure that, for every subgoal call of every thread,
-keeps track of all the answers that were already propagated.
+batched scheduling. The usage of FS design with batched scheduling,
+requires further support than with local scheduling, since when a new
+answer is found it must be immediately propagated, and we have to
+ensure that the propagation occurs on all subgoal calls once and only
+once. To do so, next we show a procedure that, for every subgoal call
+of every thread, keeps track of all the answers that were already
+propagated.
 
@@
 supporting the implementation of the PAC procedure during the
-evaluation of a tabled subgoal call $i$ using the FS design. The FS
-design, uses a subgoal entry data structure to store common
-information for a subgoal call and a subgoal frame (\emph{SF}) data
-structure to store private information about the execution of each
-thread. The PAC procedure works at the subgoal frame level, which is
+evaluation of a tabled subgoal call $i$ using the FS design. %% The FS
+%% design, uses a subgoal entry data structure to store common
+%% information for a subgoal call and a subgoal frame (\emph{SF}) data
+%% structure to store private information about the execution of each
+%% thread.
+The PAC procedure works at the subgoal frame level, which is
 private to each thread.
@@
 only showing the leaf nodes nodes $LN_1$, $LN_2$ and $LN_3$, in the
-figure. With the PAC procedure, the leaf nodes are not chained in the
-answer trie data structure. Now, the chaining process is done
-privately, and for that, we use the subgoal frame structure of each
-thread. On the subgoal frame structure we added a new field, called
+figure. %% With the PAC procedure, the leaf nodes are not chained in the
+%% answer trie data structure. Now, the chaining process is done
+%% privately, and for that, we use the subgoal frame structure of each
+%% thread.
+On the subgoal frame structure we added a new field, called
 \emph{consumer}, to store the answers found within the execution of
@@
 each thread, it does not require any of the tools that were necessary
-to support concurrency. In particular, on each hash trie level, we
-have removed the tools necessary to support concurrency, such as
-useless pointers and compare-and-swap operations. We have chosen this
-hashing mechanism, because it showed a good balance between lookup and
-insert operations~\cite{Areias-ijpp15}, but the major reason was
-mostly because of the integration in the TabMalloc memory
-allocator~\cite{Areias-12b}.
+to support concurrency, thus we have removed them from the mechanism.
+
+%% In particular, on each hash trie level, we
+%% have removed the tools necessary to support concurrency, such as
+%% useless pointers and compare-and-swap operations. We have chosen this
+%% hashing mechanism, because it showed a good balance between lookup and
+%% insert operations~\cite{Areias-ijpp15}, but the major reason was
+%% mostly because of the integration in the TabMalloc memory
+%% allocator~\cite{Areias-12b}.
 
@@
 Figure~\ref{fig_tabtries_pcc}(b) shows the state of the subgoal call
-after completion (recall that after completion of a subgoal call, the
-threads use loader nodes to consume the answers). When a thread $T$
-completes a subgoal call, it frees its private consumer structures,
-but before doing that, it checks whether another thread as already
-marked the subgoal as completed. If no other thread has done that,
-then thread $T$ not only follows its private chaining mechanism as it
-would for freeing its private nodes, but also, follows the pointers to
-the answer trie leaf nodes in order to reproduce the chain inside the
-answer trie. Since this procedure is done inside a critical region, no
-more than one thread can be doing this chaining process. Thus, in
-Figure~\ref{fig_tabtries_pcc}(b), we are showing a situation where the
-subgoal call is completed and both threads $T_1$ and $T_{k-1}$ have
-already removed their consumer answer structures and chained the leaf
-nodes inside the answer trie.
+after completion. When a thread $T$ completes a subgoal call, it frees
+its private consumer structures, but before doing that, it checks
+whether another thread as already marked the subgoal as completed. If
+no other thread has done that, then thread $T$ not only follows its
+private chaining mechanism as it would for freeing its private nodes,
+but also, follows the pointers to the answer trie leaf nodes in order
+to create a chain inside the answer trie. Since this procedure is done
+inside a critical region, no more than one thread can be doing this
+chaining process. Thus, in Figure~\ref{fig_tabtries_pcc}(b), we are
+showing a situation where the subgoal call is completed and both
+threads $T_1$ and $T_{k-1}$ have already removed their consumer answer
+structures and chained the leaf nodes inside the answer trie.
 
@@
 
-At the implementation level, the major difference between local and
-batched scheduling is in the tabling operation \emph{tabled new
-  answer}, where we decide what to do when an answer is found during
-the evaluation. This operation checks whether a newly found answer is
+At the tabling engine, the major difference between local and batched
+scheduling is in the tabling operation \emph{tabled new answer}, where
+we decide what to do when an answer is found during the
+evaluation. This operation checks whether a newly found answer is
 already in the corresponding answer trie structure and, if not,
-inserts it. This requires minor changes to the \emph{tabled new
-  answer} tabling
-operation. Algorithm~\ref{alg_table_new_answer_batched} shows how we
-have extended the tabled new answer operation to support the FS design
-with batched scheduling.
+inserts it. Algorithm~\ref{alg_table_new_answer_batched} shows how we
+have extended this operation to support FS design with batched
+scheduling.
 
@@
 We now present experimental results about the usage of the batched
-scheduling on the \emph{NS}, \emph{SS} and \emph{FS} designs. The
-environment for our experiments was a machine with 32-Core AMD Opteron
-(TM) Processor 6274 (2 sockets with 16 cores each) with 32G of main
-memory, each processor with caches $L1$, $L2$ and $L3$ respectively
-with the sizes of 64K, 2048K and 6144K, running the Linux kernel is
-the 3.16.7-200.fc20.x86\_64, with Yap 6.3 compiled with gcc 4.8.
+scheduling on FS design. To put results in perspective, we will be
+showing also the results for the NS design. The environment for our
+experiments was a machine with 32-Core AMD Opteron (TM) Processor 6274
+(2 sockets with 16 cores each) with 32G of main memory, running the
+Linux kernel is the 3.16.7-200.fc20.x86\_64 with Yap 6.3.
 
-We used five sets of benchmarks. The \emph{Large Joins} and
-\emph{WordNet} sets were obtained from the OpenRuleBench
-project~\cite{Liang-09}; the \emph{Model Checking} set includes three
-different specifications and transition relation graphs usually used
-in model checking applications; the \emph{Path Left} and \emph{Path
-  Right} sets implement two recursive definitions of the well-known
-$path/2$ predicate, that computes the transitive closure in a graph,
-using several different configurations of $edge/2$ facts. We
-experimented graphs with the \emph{BTree} configuration with depth 17,
-the \emph{Pyramid} and \emph{Cycle} configurations with depth 2000 and
-the \emph{Grid} configuration with depth 35.
-
-%% In order to have a deeper insight on the behavior of each benchmark,
-%% and therefore clarify some of the results that are presented next, we
-%% first characterize the benchmarks. The columns in
-%% Table~\ref{tab_benchs} have the following meaning: (i) {\bf calls:} is
-%% the number of different calls to tabled subgoals. It corresponds to
-%% the number of paths in the subgoal tries. (ii) {\bf trie nodes:} is
-%% the total number of trie nodes allocated in the corresponding
-%% subgoal/answer trie structures. (iii) {\bf trie depth:} is the
-%% minimum/average/maximum number of trie node levels required to
-%% represent a path in the corresponding subgoal/answer trie
-%% structures. Trie structures with smaller average values are more
-%% amenable to higher lock contention. (iv) {\bf unique:} is the number
-%% of different tabled answers found. It corresponds to the number of
-%% paths in the answer tries. (v) {\bf repeated:} is the number of
-%% redundant tabled answers found. (vi) {\bf NS:} is the average
-%% execution time, in seconds, of ten runs for 1 thread with the NS
-%% design without the TabMalloc memory allocator. In what follows, we
-%% will use these times as the base times when computing the overhead
-%% ratios for the other designs.
-
-%% By observing Table~\ref{tab_benchs}, the \emph{Mondial} benchmark,
-%% from the \emph{Large Joins} set, and the three \emph{Model
-%%   Checking} benchmarks seem to be the benchmarks least amenable to
-%% lock contention since they are the ones that find less unique answers
-%% and that have the deepest trie structures. In this regard, the
-%% \emph{Path Left} and \emph{Path Right} sets correspond to the
-%% opposite case. They find a huge number of answers and have very
-%% shallow trie structures. On the other hand, the \emph{WordNet} and
-%% \emph{Path Right} sets have the benchmarks with the largest number
-%% of different subgoal calls, which can reduce the probability of lock
-%% contention because answers can be found for different subgoal calls
-%% and therefore be inserted with minimum overlap. On the opposite side
-%% are the \emph{Join2} benchmark, from the \emph{Large Joins} set,
-%% and the \emph{Path Left} benchmarks, which have only a single tabled
-%% subgoal call.
-
-%% \begin{table}[ht]
-%% \centering
-%% \caption{Characteristics of the benchmark programs}
-%% \scalebox{0.82} { 
-%% \begin{tabular}{lrrccrrrcc}
-%% \hline\hline
-%% \multirow{2}{*}{\bf Bench}
-%% & \multicolumn{3}{c}{\bf Tabled Subgoals}
-%% &
-%% & \multicolumn{4}{c}{\bf Tabled Answers}
-%% & \multicolumn{1}{c}{\bf Time (sec)} \\ \cline{2-4}\cline{6-9}
-%% & \multicolumn{1}{c}{\bf calls}
-%% & \multicolumn{1}{c}{\bf trie nodes}
-%% & \multicolumn{1}{c}{\bf trie depth}
-%% &
-%% & \multicolumn{1}{c}{\bf unique}
-%% & \multicolumn{1}{c}{\bf repeated}
-%% & \multicolumn{1}{c}{\bf trie nodes}
-%% & \multicolumn{1}{c}{\bf trie depth}
-%% & \multicolumn{1}{c}{\bf NS} \\
-%% \hline
-%% \multicolumn{10}{l}{\bf Large Joins} \\
-%% \bf{Join2}    &       1 &       6 & 5/5/5 && 2,476,099 &         0 &
-%% 2,613,660 &    5/5/5 &  2.85 \\
-%% \bf{Mondial}  &      35 &      42 & 3/4/4 &&     2,664 & 2,452,890 &
-%%    14,334 &    6/7/7 &  0.84 \\
-%% \hline
-%% \multicolumn{10}{l}{\bf WordNet} \\
-%% \bf{Clusters} & 117,659 & 235,319 & 2/2/2 &&   166,877 &   161,853 &
-%% 284,536 &    1/1/1 &   0.83 \\
-%% \bf{Holo}     & 117,657 & 235,315 & 2/2/2 &&    74,838 &        54 &
-%% 192,495 &    1/1/1 &   0.75 \\
-%% \bf{Hyper}    & 117,657 & 235,315 & 2/2/2 &&   698,472 &     8,658 &
-%% 816,129 &    1/1/1 &   1.42 \\
-%% \bf{Hypo}     & 117,657 & 117,659 & 2/2/2 &&   698,472 &    20,341 &
-%% 816,129 &    1/1/1 &   1.53 \\
-%% \bf{Mero}     & 117,657 & 117,659 & 2/2/2 &&    74,838 &        13 &
-%% 192,495 &    1/1/1 &   0.74 \\
-%% \bf{Tropo}    & 117,657 & 235,315 & 2/2/2 &&       472 &         0 &
-%% 118,129 &    1/1/1 &   0.66 \\
-%% \hline
-%% \multicolumn{10}{l}{\bf Model Checking} \\
-%% \bf{IProto}   &       1 &       6 & 5/5/5 &&   134,361 &   385,423 &
-%% 1,554,896 &  4/51/67 & 2.70 \\
-%% \bf{Leader}   &       1 &       5 & 4/4/4 &&     1,728 &   574,786 &
-%%    41,788 & 15/80/97 & 3.51  \\
-%% \bf{Sieve}    &       1 &       7 & 6/6/6 &&       380 & 1,386,181 &
-%%     8,624 & 21/53/58 & 18.50 \\
-%% \hline
-%% \multicolumn{10}{l}{\bf Path Left} \\
-%% \bf{BTree}    &       1 &       3 & 2/2/2 && 1,966,082 &         0 &
-%% 2,031,618 &    2/2/2 &   1.53 \\
-%% \bf{Cycle}    &       1 &       3 & 2/2/2 && 4,000,000 &     2,000 &
-%% 4,002,001 &    2/2/2 &   3.52 \\
-%% \bf{Grid}     &       1 &       3 & 2/2/2 && 1,500,625 & 4,335,135 &
-%% 1,501,851 &    2/2/2 &  1.93 \\
-%% \bf{Pyramid}  &       1 &       3 & 2/2/2 && 3,374,250 & 1,124,250 &
-%% 3,377,250 &    2/2/2 &   3.08 \\
-%% \hline
-%% \multicolumn{10}{l}{\bf Path Right} \\
-%% \bf{BTree}    & 131,071 & 262,143 & 2/2/2 && 3,801,094 &         0 &
-%% 3,997,700 &    1/2/2 &  2.33 \\
-%% \bf{Cycle}    &   2,001 &   4,003 & 2/2/2 && 8,000,000 &     4,000 &
-%% 8,004,001 &    1/2/2 &  3.55 \\
-%% \bf{Grid}     &   1,226 &   2,453 & 2/2/2 && 3,001,250 & 8,670,270 &
-%% 3,003,701 &    1/2/2 &  2.32 \\
-%% \bf{Pyramid}  &   3,000 &   6,001 & 2/2/2 && 6,745,501 & 2,247,001 &
-%% 6,751,500 &    1/2/2 &  3.17 \\
-%% \hline\hline
-%% \end{tabular}}
-%% \label{tab_benchs}
-%% \end{table}
-
-%% Note that for the moment our goal is to evaluate the robustness of our
-%% implementation when exposed to \emph{worst case} scenarios. By
-%% focusing on \emph{worst case} scenarios, we show the lowest bounds in
-%% terms of performance that each design might achieve when applied/used
-%% with other real world applications/programs. Moreover, by testing the
-%% framework with worst case scenarios, we avoid the peculiarities of the
-%% program at hand and we try to focus on measuring the real value of our
-%% designs. Thus, we will follow a common approach to create worst case
-%% scenarios and we will run all threads starting with the same query
-%% goal. By doing this, it is expected that all threads will access the
-%% table space, to check/insert for subgoals and answers, at similar
-%% times, thus causing a huge stress on the same critical regions. To put
-%% the results in perspective, we experimented with $1$, $8$, $16$, $24$
-%% and $32$ threads (the maximum number of cores available in our
-%% machine) with local and batched scheduling on the five sets of
-%% benchmarks presented in the early subsection.
+For the experimentation, we used our memory allocator described in the
+work~\cite{Areias-12b}, later named \emph{TabMalloc}, and we the same
+five sets of benchmarks. These benchmarks create \emph{worst case}
+scenarios, where we are able to show the lowest bounds in terms of
+performance that each design might achieve when applied/used with
+other real world applications/programs.
 
@@
 compared with the NS design with 1 thread (running with local
-scheduling), for the NS and FS designs (all running TabMalloc), when
-running 1, 8, 16, 24 and 32 threads with local and batched scheduling
-on the five sets of benchmarks. For each design, the table has then
-two columns, a column with {\bf Local} that shows results for the
-local scheduling and a column with {\bf Batched} with results for
-batched scheduling. The overhead results presented in both {\bf Local}
-and {\bf Batched} columns use as base time the execution times
-presented in the {\bf NS} column of the Table~\ref{tab_benchs}.
+scheduling without TabMalloc), for the NS and FS designs, when running
+1, 8, 16, 24 and 32 threads, with TabMalloc and, local scheduling
+(column \emph{Local}) and batched scheduling (column \emph{Batched})
+strategies. In order to give a fair weight to each benchmark, the
+overhead ratio is calculated as follows. We begin by running ten times
+each benchmark $B$ for each design $D$ with $T$ threads. Then, we
+calculate the average of those ten runs and use that value ($D_{BT}$)
+to put it in perspective against the base time, which is the average
+of the ten runs of the NS design with $1$ thread ($NS_{B1}$). For
+that, we use the following formula for the overhead $O_{DBT} = D_{BT}
+/ NS_{B1}$.  After calculating all the overheads $O_{DBT}$ for a
+certain design $D$ and number of threads $T$ corresponding to the
+several benchmarks $B$, we calculate the respective minimum, average,
+maximum and standard deviation overhead ratios. Thus, the lowest ratio
+is the one that represents lowest overhead and it is the best ratio.
 
 \setlength{\tabcolsep}{12pt}
-
 \begin{table}[!ht]
@@
 \centering
-\caption{Overhead ratios, when compared with the NS design with $1$
-  thread (running with local scheduling without TabMalloc) for the NS
-  and FS designs (with TabMalloc), when running 1, 8, 16, 24 and 32
-  threads with local and batched scheduling on the five sets of
-  benchmarks (best ratios by row and by design for the Minimum,
-  Average and Maximum are in bold)}
+%% \caption{Overhead ratios, when compared with the NS design with $1$
+%%   thread (running with local scheduling without TabMalloc) for the NS
+%%   and FS designs (with TabMalloc), when running 1, 8, 16, 24 and 32
+%%   threads with local and batched scheduling on the five sets of
+%%   benchmarks (best ratios by row and by design for the Minimum,
+%%   Average and Maximum are in bold)}
+\caption{Overhead ratios, when compared with the NS design with 1
+  thread for the NS and FS designs, when running 1, 8, 16, 24 and 32
+  threads with local and batched scheduling (lowest is the best and it
+  is marked in bold by row and by design for the Minimum, Average and
+  Maximum)}
 \vspace{-0.5\intextsep}

### 2015-04-16 paper.tex
conclusions

@@
 several benchmarks $B$, we calculate the respective minimum, average,
-maximum and standard deviation overhead ratios. Thus, the lowest ratio
-is the one that represents lowest overhead and it is the best ratio.
+maximum and standard deviation overhead ratios.
 
@@
 
-By observing Table~\ref{tab_batched_overhead}, we can see that, for
-one thread, on average, local scheduling is sightly better than
-batched on the two designs. For the NS design we have $0.78$ and
-$0.82$ and for the FS design we have $1.30$ and $1.46$ average
-overhead ratios, for the local and batched scheduling strategies,
-respectively. As we scale the number of threads, one can observe that,
-for the NS design both scheduling strategies have similar minimum,
-average and maximum overhead ratios. For the FS design, the best
-minimum overhead ratio on $1$, $8$, $16$, $24$ and $32$ threads is
-always the batched scheduling. For the average and maximum overhead
-ratio, local scheduling is always better than batched
-scheduling. However, for the average ratio, the results between both
-strategies is quite close. We consider this to be quite significant,
-once the FS design with batched scheduling requires the support of the
-PAC procedure. For the maximum overhead ratio, the difference between
-both strategies is higher than for the average overhead ratio, however
-we hope to reduce the difference with further improvements in the PAC
-procedure.
+By observing Table~\ref{tab_batched_overhead}, we can see that, for 1
+thread, on the average overhead ratio, local scheduling is sightly
+better than batched on the two designs. For the NS design we have
+$0.78$ and $0.82$ and for the FS design we have $1.30$ and $1.46$
+average overhead ratios, for the local and batched scheduling
+strategies, respectively.
+
+%%  As we scale the number of threads, one can observe that,
+%% for the NS design both scheduling strategies have similar minimum,
+%% average and maximum overhead ratios. For the FS design, the best
+%% minimum overhead ratio on $1$, $8$, $16$, $24$ and $32$ threads is
+%% always the batched scheduling. For the average and maximum overhead
+%% ratio, local scheduling is always better than batched
+%% scheduling. However, for the average ratio, the results between both
+%% strategies is quite close. We consider this to be quite significant,
+%% once the FS design with batched scheduling requires the support of the
+%% PAC procedure. For the maximum overhead ratio, the difference between
+%% both strategies is higher than for the average overhead ratio, however
+%% we hope to reduce the difference with further improvements in the PAC
+%% procedure.
 
@@
 \section{Conclusions and Further Work}
-.........
+We have presented a simple and novel approach for combining the FS
+design with batched scheduling. Experimental results showed that one
+average, the extra structures required for the combination do not seem
+to have a big impact in the performance. Further work will include
+experimentation in more applications and improvements in the PAC
+procedure.
 

### 2015-04-16 paper.tex
removing stuff

@@
 By observing Table~\ref{tab_batched_overhead}, we can see that, for 1
-thread, on the average overhead ratio, local scheduling is sightly
-better than batched on the two designs. For the NS design we have
-$0.78$ and $0.82$ and for the FS design we have $1.30$ and $1.46$
-average overhead ratios, for the local and batched scheduling
-strategies, respectively.
+thread, on the minimum overhead ratio, local scheduling is better on
+the NS design, while batched scheduling is better in the FS
+design. For the average and maximum overhead ratios, in the FS design,
+the best strategy is always the local scheduling.
 

### 2015-04-16 paper.tex
removing stuff

@@
 
+As we scale the number of threads, for the NS design both scheduling
+strategies have similar minimum, average and maximum overhead
+ratios. For the FS design, the best minimum overhead ratio is always
+for the batched scheduling. For the average and maximum overhead
+ratio, local scheduling is always better than batched
+scheduling. However, for the average ratio, the results between both
+strategies is quite close. For the maximum overhead ratio, the
+difference between both strategies is significantly higher.
+
 %%  As we scale the number of threads, one can observe that,
@@
 We have presented a simple and novel approach for combining the FS
-design with batched scheduling. Experimental results showed that one
+design with batched scheduling. Experimental results showed that on
 average, the extra structures required for the combination do not seem
 to have a big impact in the performance. Further work will include
-experimentation in more applications and improvements in the PAC
-procedure.
+experimentation in other real world applications.
 

### 2015-04-16 paper.tex
removing stuff
abstract - ok

@@
 \newcommand{\LINECOMMENT}[1]{$\triangleright$ #1}
-
 %\setlength{\belowcaptionskip}{-26pt}
@@
 
-\title{On Extending a Full-Sharing Design with Batched Scheduling}
+\title{On Extending a Full-Sharing Multithreaded Tabling Design with
+  Batched Scheduling}
 
@@
 \begin{abstract}
+Tabling is an implementation technique that overcomes some limitations
+of traditional Prolog systems in dealing with redundant
+sub-computations and recursion. Multithreading in Prolog is the
+ability to concurrently perform computations, in which each
+computation runs independently but shares the program clauses. When
+multithreading is combined with tabling, we have the best of both
+worlds, since we can exploit the combination of higher procedural
+control with higher declarative semantics.  To support this
+combination, the Yap system has at engine level, multiple designs that
+vary from a No-Sharing design, where each thread allocates fully
+private tables for each new tabled subgoal called during its
+computation, to a Full-Sharing design, where threads share the
+complete table space. Arguably, batched scheduling is one of the most
+successful tabling scheduling strategies. In this work, we propose an
+extension of the Full-Sharing design to support batched
+scheduling. Experimental results show that, despite the extra data
+structures required to support the extension, the FS design remains
+quite competitive.
 
-  \textbf{Keywords:} Multithreading, Tabling, Concurrency, Batched
-  Scheduling
+\textbf{Keywords:} Logic Programming, Multithreading, Tabling, Scheduling.
 \end{abstract}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Introduction}
@@
 redundant sub-computations and recursion. Tabling consists of storing
-intermediate answers for subgoals so that they can be reused when a
-repeated subgoal appears during the resolution process. Tabling has
-become a popular and successful technique thanks to the
-ground-breaking work in the XSB Prolog system and in particular in the
-SLG-WAM engine~\cite{Sagonas-98}, the most successful engine of
-XSB. The success of SLG-WAM led to several alternative implementations
-that differ in the execution rule, in the data-structures used to
-implement tabling, and in the changes to the underlying Prolog
+intermediate answers for subgoals in a proper data structure, called
+the \emph{table space}, so that they can be reused when a repeated
+subgoal appears during the resolution process. Tabling has become a
+popular and successful technique thanks to the ground-breaking work in
+the XSB Prolog system and in particular in the SLG-WAM
+engine~\cite{Sagonas-98}, the most successful engine of XSB. The
+success of SLG-WAM led to several alternative implementations that
+differ in the execution rule, in the data-structures used to implement
+tabling, and in the changes to the underlying Prolog
 engine. Implementations of tabling are now widely available in systems
@@
 best of our knowledge, XSB~\cite{Marques-08} and Yap~\cite{Areias-12a}
-were the only Prolog systems that were able to combine both
+are the only Prolog systems that support the combination of
 multithreading with tabling. Yap implements a SWI-Prolog compatible
-multi-threading library~\cite{Wielemaker-03}. Yap's threads have their
+multithreading library~\cite{Wielemaker-03}. Yap's threads have their
 own execution stacks and only share the code area where predicates,
@@
 tabled evaluation, a thread views its tables as private but, at the
-engine level, Yap has three designs that vary from a \emph{No-Sharing}
-(NS) design, where each thread allocates fully private tables for each
-new tabled subgoal called during its computation, to a
-\emph{Full-Sharing} (FS), where threads share the complete table
-space.
+engine level, Yap has three designs~\cite{Areias-12a} that vary from a
+\emph{No-Sharing} (NS) design, where each thread allocates fully
+private tables for each new tabled subgoal called during its
+computation, to a \emph{Full-Sharing} (FS) design, where threads share
+the complete table space.
 
-In this paper we discuss a novel approach for combining the FS design
-with batched scheduling and we present a performance analysis
-comparison between local scheduling with batched
-scheduling. Experimental results show that despite the extra data
-structures required to combine FS design with batched scheduling, the
-execution time of combination is still quite competitive when
-comparing against the FS with local scheduling.
+The decision about the evaluation flow is determined by the
+\emph{scheduling strategy}. Different strategies may have a
+significant impact on performance, and may lead to a different
+ordering of solutions to the query goal. Arguably, the two most
+successful tabling scheduling strategies are \emph{local scheduling}
+and \emph{batched scheduling}~\cite{Freire-96}. Local scheduling tries
+to complete subgoals as soon as possible. When new answers are found,
+they are added to the table space and the evaluation fails. Answers
+are only returned when all program clauses for the subgoal at hand
+were resolved. Batched scheduling favors forward execution first,
+backtracking next, and consuming answers or completion last. It thus
+tries to delay the need to move around the search tree by batching the
+return of answers to repeated subgoals. When new answers are found for
+a particular tabled subgoal, they are added to the table space and the
+evaluation continues.
+
+With the FS design, all tables are shared. Thus, since several threads
+can be inserting answers in the same table, when an answer already
+exists, it is not possible to determine if the answer is new or
+repeated for a certain thread without further support. However, for
+local scheduling this is not a problem because, for repeated and new
+answers, local scheduling always fails. The problem is with batched
+scheduling that requires further support to work with the FS design
+since, with batched scheduling, only repeated answers should fail. We
+have thus to detect when a new found answer is also a new answer for
+the thread at hand in order to ensure that new answers for a given
+subgoal call are correctly propagated one and only once to all
+thread's calls.
+
+In this work, we propose an extension to the table space data
+structures, which we named \emph{Private Answer Chaining (PAC)}, as a
+way to keep track, per thread and subgoal call, of the answers that
+were already found and propagated to all repeated calls. We discuss in
+detail our proposal for extending the FS design with batched
+scheduling and we present a performance analysis comparison between
+local and batched scheduling. Experimental results show that, despite
+the extra PAC data structures required to support batched scheduling
+with the FS design, the execution time of the combination is still
+quite competitive when comparing against the FS design with local
+scheduling.
 
@@
 introduce some background and related work. Then, we describe our
-approach for combining batched scheduling with the FS design. Next, we
-show the most important implementation details and finally, we discuss
+approach for combining batched scheduling with the FS design and we
+discuss the most important implementation details. Finally, we present
 experimental results and we end by outlining some conclusions.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Background}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{Table Space Organization}
@@
 \begin{wrapfigure}{r}{4.5cm}
-\vspace{-\intextsep}
+%\vspace{-\intextsep}
 \centering
@@
 \label{fig_table_space}
-\vspace{-0.5\intextsep}
+%\vspace{-0.5\intextsep}
 \end{wrapfigure}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{Yap's Multithreaded Tabling Support}
 
-Yap implements a SWI-Prolog compatible multi-threading
-library~\cite{Wielemaker-03}. Like in SWI-Prolog, Yap's threads have
-their own execution stacks and only share the code area where
-predicates, records, flags and other global non-backtrackable data are
-stored.
-
-In a multithreaded tabling evaluation, a thread views its tables as
-private but, at the engine level, Yap has three designs that vary from
-a \emph{No-Sharing} (NS) design, where each thread allocates fully
-private tables for each new tabled subgoal called during its
-computation, to a \emph{Full-Sharing} (FS), where threads share the
-complete table space. Figure~\ref{fig_table_space} shows Yap's
-multithreaded tabling support with an abstraction layer to which
-threads interact (the threads represented in the bottom of the
-figure), and at the engine level the configuration of the NS and FS
-designs in a situation where several different threads evaluated the
-same tabled subgoal call $i$.
-
-In the NS design, each thread allocates fully private tables for each
-new subgoal called during its computation. One can observe that the
-table entry data structure still stores the common information for the
-predicate (such as the arity or the evaluation strategy), and then
-each thread $t$ has its own cell $T_t$ inside a \emph{bucket array}
-which points to the private data structures.
+In Yap, a thread views its tables as private but, at the engine level,
+it uses a common table space where tables are shared among all
+threads. Yap implements three designs for concurrent tabling support
+that vary from a \emph{No-Sharing} (NS) design, where each thread
+allocates fully private tables for each new tabled subgoal called
+during its computation, to a \emph{Full-Sharing} (FS) design, where
+threads share the complete table space. Figure~\ref{fig_table_space}
+shows Yap's multithreaded table space organization for the NS and FS
+designs, where an interface layer abstracts the design being used at
+the engine level. The figure illustrates the main differences between
+the two designs for a situation where several threads are evaluating
+the same tabled subgoal call $call\_i$.
 
 \begin{figure}[!ht]
-\vspace{-\intextsep}
+%\vspace{-0.5\intextsep}
 \centering
 \includegraphics[width=9.5cm]{figures/yap-mt.pdf}
-\caption{Yap's multithreaded tabled space using the NS and FS designs}
+\caption{Yap's multithreaded table space organization for the NS and FS designs}
 \label{fig_yap_mt_support}
-\vspace{-1.0\intextsep}
+%\vspace{-1.0\intextsep}
 \end{figure}
 
-In the FS design, part of the subgoal frame information (the
-\emph{subgoal entry} data structure) and the answer trie structure are
-now shared among all threads. The previous subgoal frame data
-structure was split into two: the \emph{subgoal entry} stores common
-information for the subgoal call (such as the pointer to the shared
-answer trie structure); the remaining information (the \emph{subgoal
-  frame} data structure) remains private to each thread.
+When using the NS design, each thread allocates fully private tables
+for each new subgoal called during its computation. One can observe
+that the table entry data structure still stores the common
+information for the predicate (such as the arity or the evaluation
+strategy), and then each thread $t$ has its own cell $T_t$ inside a
+\emph{bucket array} which points to the private data structures.
+
+When using the FS design, the subgoal and answer trie structures and
+part of the subgoal frame information (the \emph{subgoal entry} data
+structure in Fig.~\ref{fig_yap_mt_support}) are shared among all
+threads. The previous subgoal frame data structure was split in two:
+the \emph{subgoal entry} stores common information for the subgoal
+call (such as the pointer to the shared answer trie structure); the
+remaining information is kept private to each thread in the
+\emph{subgoal frame} data structure.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{Scheduling Strategies}
 
-The decision about the evaluation flow is determined by the
-\emph{scheduling strategy}. Different strategies may have a
-significant impact on performance, and may lead to a different
-ordering of solutions to the query goal. Arguably, the two most
-successful tabling scheduling strategies are \emph{local scheduling}
-and \emph{batched scheduling}~\cite{Freire-96}.
+Local scheduling schedules the evaluation of a program in a
+breath-first manner. It favors backtracking first with completion
+instead of forward execution, leaving the consumption of answers for
+last. Local scheduling only allows a \emph{Cluster of Dependent
+  Subgoals} (CDS) to return answers after a fixpoint point has been
+reached~\cite{Freire-96}.
 
-Local scheduling strategy schedules the evaluation of a program in a
-breath-first manner. It favors the backtracking first with completion
-instead of the forward execution, leaving the consumption of answers
-for last. Thus, it only allows a Cluster of Dependent Subgoals (CDS)
-to return answers only after the completion point has been
-reached~\cite{Freire-96}. %% In other words, the local scheduling tries
-%% to keep a CDS as minimal as possible. When new answers are found, they
-%% are added to the table space and the computation fails as consequence,
-%% tabled subgoals inside a CDS propagate their answers to outside the
-%% CDS only after its completion point is found.
+In other words, local scheduling tries
+to keep CDS as minimal as possible. When new answers are found, they
+are added to the table space and the computation fails as consequence,
+tabled subgoals inside a CDS propagate their answers to outside the
+CDS only after its completion point is found.
+
 Local scheduling causes a sooner completion of subgoals, which creates
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Extending Full-Sharing with Batched Scheduling}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{Our Approach}
@@
 \begin{figure}[!ht]
-\vspace{-\intextsep}
+%\vspace{-\intextsep}
 \centering
 \includegraphics[width=10.5cm]{figures/pcc.pdf}
-\caption{The \emph{FS} design with the PAC procedure - (a) private
-  chaining and (b) public chaining}
+\caption{The PAC procedure using (a) private chaining and (b) public
+  chaining}
 \label{fig_tabtries_pcc}
-\vspace{-1.0\intextsep}
+%\vspace{-1.0\intextsep}
 \end{figure}
@@
 \begin{table}[!ht]
-\vspace{-\intextsep}
+%\vspace{-\intextsep}
 \centering
@@
   Maximum)}
-\vspace{-0.5\intextsep}
+%\vspace{-0.5\intextsep}
 %\scalebox{1.00} {
@@
 \label{tab_batched_overhead}
-\vspace{-0.7\intextsep}
+%\vspace{-0.7\intextsep}
 \end{table}

### 2015-04-17 paper.tex
final version

@@
 control with higher declarative semantics. To support this
-combination, the Yap system has at engine level, multiple designs that
-vary from a No-Sharing design, where each thread allocates fully
-private tables for each new tabled subgoal called during its
-computation, to a Full-Sharing design, where threads share the
-complete table space. Arguably, batched scheduling is one of the most
-successful tabling scheduling strategies. In this work, we propose an
-extension of the Full-Sharing design to support batched
+combination, the Yap Prolog system has, at engine level, multiple
+designs that vary from a No-Sharing design, where each thread
+allocates fully private tables for each new tabled subgoal called
+during its computation, to a Full-Sharing design, where threads share
+the complete table space. Arguably, batched scheduling is one of the
+most successful tabling scheduling strategies. In this work, we
+propose an extension of the Full-Sharing design to support batched
 scheduling. Experimental results show that, despite the extra data
 structures required to support the extension, the FS design remains
-quite competitive.
+quite competitive.\\
 
@@
 since, with batched scheduling, only repeated answers should fail. We
-have thus to detect when a new found answer is also a new answer for
+have thus to detect when a newly found answer is also a new answer for
 the thread at hand in order to ensure that new answers for a given
 subgoal call are correctly propagated one and only once to all
-thread's calls.
+thread's repeated calls.
 
@@
 In Yap, a thread views its tables as private but, at the engine level,
-it uses a common table space where tables are shared among all
-threads. Yap implements three designs for concurrent tabling support
-that vary from a \emph{No-Sharing} (NS) design, where each thread
-allocates fully private tables for each new tabled subgoal called
-during its computation, to a \emph{Full-Sharing} (FS) design, where
-threads share the complete table space. Figure~\ref{fig_table_space}
-shows Yap's multithreaded table space organization for the NS and FS
-designs, where an interface layer abstracts the design being used at
-the engine level. The figure illustrates the main differences between
-the two designs for a situation where several threads are evaluating
-the same tabled subgoal call $call\_i$.
+it implements three designs for concurrent tabling support that vary
+from a \emph{No-Sharing} (NS) design, where each thread allocates
+fully private tables for each new tabled subgoal called during its
+computation, to a \emph{Full-Sharing} (FS) design, where threads share
+the complete table space. Figure~\ref{fig_table_space} shows Yap's
+multithreaded table space organization for the NS and FS designs,
+where an interface layer abstracts the design being used at the engine
+level. The figure illustrates the main differences between the two
+designs for a situation where several threads are evaluating the same
+tabled subgoal call $call\_i$.
 
@@
 last. Local scheduling only allows a \emph{Cluster of Dependent
-  Subgoals} (CDS) to return answers after a fixpoint point has been
-reached~\cite{Freire-96}.
-
-In other words, local scheduling tries
-to keep CDS as minimal as possible. When new answers are found, they
-are added to the table space and the computation fails as consequence,
-tabled subgoals inside a CDS propagate their answers to outside the
-CDS only after its completion point is found.
-
-Local scheduling causes a sooner completion of subgoals, which creates
-less complex dependencies between them.
+  Subgoals} (CDS) to return answers after a fix-point has been
+reached~\cite{Freire-96}. In other words, local scheduling tries to
+keep a CDS as minimal as possible, thus creating less complex
+dependencies between subgoals, which causes a sooner completion of
+subgoals.
 
@@
 in creating dependencies to older subgoals, therefore enlarging the
-current CDS~\cite{Sagonas-98} and delaying the completion point to an
-older generator node.
+current CDS and delaying the fix-point that guarantees that all
+dependent subgoals in a CDS are completely
+evaluated~\cite{Sagonas-98}.
 
@@
 
-In this section, we show how we have extended the FS design to support
-batched scheduling. The usage of FS design with batched scheduling,
-requires further support than with local scheduling, since when a new
-answer is found it must be immediately propagated, and we have to
-ensure that the propagation occurs on all subgoal calls once and only
-once. To do so, next we show a procedure that, for every subgoal call
-of every thread, keeps track of all the answers that were already
-propagated.
+In this section, we describe our proposal to support the combination
+of batched scheduling with the FS design. Remember that batched
+scheduling requires further support to work with the FS design
+because, with batched scheduling, we have to detect when a newly found
+answer is also a new answer for the thread at hand in order to ensure
+that new answers for a given subgoal call are correctly propagated one
+and only once to all thread’s repeated calls. To solve that, we propose
+an extension to the table space data structures, which we named
+\emph{Private Answer Chaining} (PAC), as a way to keep track, per
+thread and subgoal call, of the answers that were already found and
+propagated to the thread's repeated calls.
 
@@
 
-The key idea of our approach, which we named Privately-consumed Answer
-Chaining (PAC), is then to extend the \emph{FS} design with batched
-scheduling, by chaining privately for each subgoal call, the answers
-that were already consumed by a thread. Since the procedure is
-private, it will only affect the thread that is doing it. At the end,
-when the evaluation is complete, i.e, when a subgoal call is marked as
-complete, we put one of the private chain as public, so that from that
-point on all threads can use that chain in complete (only reading)
-mode.
-
-Figure~\ref{fig_tabtries_pcc} shows the key data structures for
-supporting the implementation of the PAC procedure during the
-evaluation of a tabled subgoal call $i$ using the FS design. %% The FS
-%% design, uses a subgoal entry data structure to store common
-%% information for a subgoal call and a subgoal frame (\emph{SF}) data
-%% structure to store private information about the execution of each
-%% thread.
-The PAC procedure works at the subgoal frame level, which is
-private to each thread.
+The PAC procedure works at the subgoal frame level, which is private
+to each thread. The key idea of the PAC data structure is then to
+extend subgoal frames with an auxiliary private chaining of answers
+for each subgoal call, in order to keep track of the answers already
+found for the call. Later, if a thread completes a subgoal's
+evaluation, i.e, if the subgoal's table is marked as complete, its PAC
+is made public, so that from that point on all threads can use that
+chain in complete (only reading) mode. Figure~\ref{fig_tabtries_pcc}
+illustrates the new data structures involved in the implementation of
+our PAC's proposal for a situation where several threads are
+evaluating the same tabled subgoal call $call\_i$.
 
@@
 \includegraphics[width=10.5cm]{figures/pcc.pdf}
-\caption{The PAC procedure using (a) private chaining and (b) public
+\caption{PAC's data structures for (a) private and (b) public
   chaining}
@@
 Figure~\ref{fig_tabtries_pcc}(a) shows then a situation where two
-threads, $T_1$ and $T_{k-1}$, are sharing the same subgoal entry call
-$i$ when the subgoal is still under evaluation, i.e., the subgoal is
-not yet complete. The current state of the evaluation shows an answer
-trie with $3$ answers found for the subgoal call $i$. For the sake of
-simplicity, we are omitting the internal answer trie nodes and we are
-only showing the leaf nodes nodes $LN_1$, $LN_2$ and $LN_3$, in the
-figure. %% With the PAC procedure, the leaf nodes are not chained in the
-%% answer trie data structure. Now, the chaining process is done
-%% privately, and for that, we use the subgoal frame structure of each
-%% thread.
-On the subgoal frame structure we added a new field, called
-\emph{consumer}, to store the answers found within the execution of
-the thread. In order to minimize the impact of the \emph{PAC}
-optimization, each node within the new consumer answer structure has
-two fields: (i) an entry pointer, which points to the corresponding
-leaf node in the answer trie data structure; (ii) a next pointer to
-chain the answers within the consumer structure. To maintain a good
-performance, when the number of nodes exceeds a certain threshold, we
-use a hash trie mechanism design similar to the one presented in the
+threads, $T_1$ and $T_{k-1}$, are sharing the same subgoal entry for
+the call $call\_i$ when the subgoal is still under evaluation, i.e.,
+the subgoal is not yet complete. The current state of the evaluation
+shows an answer trie with 3 answers found for the subgoal call
+$call\_i$. For the sake of simplicity, we are omitting the internal
+answer trie nodes and we are only showing the leaf nodes nodes $LN_1$,
+$LN_2$ and $LN_3$, in the figure.
+
+With PAC support, the leaf nodes are not chained in the answer trie
+data structure, as usual. Now, the chaining process is done privately,
+and for that, we use the subgoal frame structure of each thread. On
+the subgoal frame structure we added a new field, called
+\emph{Answers}, to store the answers found within the execution of the
+thread. In order to minimize PAC's impact, each answer node in the
+private chaining has only two fields: (i) an entry pointer, which
+points to the corresponding leaf node in the answer trie data
+structure; and (ii) a next pointer to chain the nodes in the private
+chaining. To maintain a good performance, when the number of answer
+nodes exceeds a certain threshold, we use a hash trie mechanism design
+similar to the one presented in the
 work~\cite{Areias-ijpp15}. However, since this mechanism is private to
-each thread, it does not require any of the tools that were necessary
-to support concurrency, thus we have removed them from the mechanism.
+each thread, it does not require any concurrency support.
 
-%% In particular, on each hash trie level, we
-%% have removed the tools necessary to support concurrency, such as
-%% useless pointers and compare-and-swap operations. We have chosen this
-%% hashing mechanism, because it showed a good balance between lookup and
-%% insert operations~\cite{Areias-ijpp15}, but the major reason was
-%% mostly because of the integration in the TabMalloc memory
-%% allocator~\cite{Areias-12b}.
-
-Going back to Figure~\ref{fig_tabtries_pcc}(a), the consumer answer
+Going back to Fig.~\ref{fig_tabtries_pcc}(a), the PAC's data
 structures represent then two different situations where threads can
 be evaluating a subgoal call. Thread $T_1$ has only found one answer
-and it is using a direct consumer answer chaining to access the node
+and it is using a direct answer chaining to access the node
 $LN_1$. Thread $T_{k-1}$ was already found three answers for the
-subgoal call and it is already using the hash trie mechanism within
-its consumer answer structure. The consumer nodes are chained between
-themselves, thus that consumer nodes belonging to thread $T_{k-1}$ can
-consume the answers as in the original mechanism.
+subgoal call and it is using the hash trie mechanism within its
+private chaining. Please note that the answer nodes are still chained
+between themselves, thus that repeated calls belonging to thread
+$T_{k-1}$ can consume the answers as in the original mechanism.
 
@@
 inside a critical region, no more than one thread can be doing this
-chaining process. Thus, in Figure~\ref{fig_tabtries_pcc}(b), we are
-showing a situation where the subgoal call is completed and both
-threads $T_1$ and $T_{k-1}$ have already removed their consumer answer
-structures and chained the leaf nodes inside the answer trie.
+chaining process. Thus, in Fig.~\ref{fig_tabtries_pcc}(b), we are
+showing the situation where the subgoal call is completed and both
+threads $T_1$ and $T_{k-1}$ have already chained the leaf nodes inside
+the answer trie and removed their private chaining structures.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{Implementations Details}
@@
 inserts it. Algorithm~\ref{alg_table_new_answer_batched} shows how we
-have extended this operation to support FS design with batched
+have extended this operation to support the FS design with batched
 scheduling.
@@
 
-The algorithm receives two arguments: the new answer found during the
-evaluation (\emph{ANS}) and the subgoal frame which corresponds to the
-call at hand (\emph{SF}). The algorithm begins by checking/inserting
-the given \emph{ANS} into the answer trie structure, which will return
-the leaf node for the path representing \emph{ANS} (line 1).
-
-Then, it checks/inserts the given \emph{leaf} node into the private
-consumer chain for the current thread, which will return the
-corresponding chain node (line 2). Next in line 3, it tests whether
-the chain node already existed in the consumer chain, i.e., if it was
-inserted or not by the current check/insert operation in order to
-return failure (line 4), or it proceeds with marking the answer
-\emph{ANS} has found (line 6). At the end (lines 7 to 10), it returns
-failure if local scheduling is active (line 8), otherwise, the batched
-scheduling is active, thus it propagates the answer \emph{ANS} (line
-10).
+The algorithm receives two arguments: the newly answer found during
+the evaluation (\emph{ANS}) and the subgoal frame which corresponds to
+the call at hand (\emph{SF}). The algorithm begins by
+checking/inserting the given \emph{ANS} into the answer trie
+structure, which will return the leaf node for the path representing
+\emph{ANS} (line 1). Then, it checks/inserts the given \emph{leaf}
+node into the private chaining for the current thread, which will
+return the corresponding answer chain node (line 2). Next in line 3,
+it tests whether the answer chain node already existed in the chain,
+i.e., if it was inserted or not by the current check/insert operation
+in order to return failure (line 4), or it proceeds with marking the
+answer \emph{ANS} has found (line 6). At the end (lines 7 to 10), it
+returns failure if local scheduling is active (line 8), otherwise,
+batched scheduling is active, and it proceeds by propagating the
+answer \emph{ANS} to the current execution environment (line 10).
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Performance Analysis}
 
-We now present experimental results about the usage of the batched
-scheduling on FS design. To put results in perspective, we will be
-showing also the results for the NS design. The environment for our
-experiments was a machine with 32-Core AMD Opteron (TM) Processor 6274
-(2 sockets with 16 cores each) with 32G of main memory, running the
-Linux kernel is the 3.16.7-200.fc20.x86\_64 with Yap 6.3.
+We now present experimental results about the usage of the FS design
+with batched scheduling. To put results in perspective, we will be
+showing also the results for local scheduling and for the NS
+design.
 
-For the experimentation, we used our memory allocator described in the
-work~\cite{Areias-12b}, later named \emph{TabMalloc}, and we the same
-five sets of benchmarks. These benchmarks create \emph{worst case}
-scenarios, where we are able to show the lowest bounds in terms of
+The environment for our experiments was a machine with 32-Core AMD
+Opteron (TM) Processor 6274 (2 sockets with 16 cores each) with 32G of
+main memory, running the Linux kernel is the 3.16.7-200.fc20.x86\_64
+with Yap Prolog 6.3. For the experiments, we used the \emph{TabMalloc}
+memory allocator~\cite{Areias-12b} and the set of benchmarks described
+in~\cite{Areias-12b}. These benchmarks create \emph{worst case
+  scenarios}, where we are able to show the lowest bounds in terms of
 performance that each design might achieve when applied/used with
@@
 \begin{table}[!ht]
-%\vspace{-\intextsep}
 \centering
-%% \caption{Overhead ratios, when compared with the NS design with $1$
-%%   thread (running with local scheduling without TabMalloc) for the NS
-%%   and FS designs (with TabMalloc), when running 1, 8, 16, 24 and 32
-%%   threads with local and batched scheduling on the five sets of
-%%   benchmarks (best ratios by row and by design for the Minimum,
-%%   Average and Maximum are in bold)}
-\caption{Overhead ratios, when compared with the NS design with 1
-  thread for the NS and FS designs, when running 1, 8, 16, 24 and 32
-  threads with local and batched scheduling (lowest is the best and it
-  is marked in bold by row and by design for the Minimum, Average and
-  Maximum)}
-%\vspace{-0.5\intextsep}
-%\scalebox{1.00} {
-\begin{tabular}{ll|cc|cc}
-\hline\hline
-\multicolumn{2}{c|}{\multirow{2}{*}{\bf Threads}} &
+\caption{Overhead ratios, when compared with the NS design with $1$
+  thread (running with local scheduling and without TabMalloc) for the NS
+  and FS designs (with TabMalloc), when running 1, 8, 16, 24 and 32
+  threads with local and batched scheduling (best ratios by row and by
+  design for the Minimum, Average and Maximum are in bold)}
+\begin{tabular}{ll||cc|cc}
+\multicolumn{2}{c||}{\multirow{2}{*}{\bf Threads}} &
 \multicolumn{2}{c|}{\multirow{1}{*}{\bf NS}} &
@@
 & \multicolumn{1}{c}{\bf Batched}\\
-\hline
+\hline\hline
 \multirow{4}{*}{\bf 1}
@@
 & {\bf StD }& 0.45& 0.45& 0.74& 1.02\\
-\hline\hline
-\end{tabular}%}
+\end{tabular}
 \label{tab_batched_overhead}
-%\vspace{-0.7\intextsep}
 \end{table}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Conclusions and Further Work}
+
 We have presented a simple and novel approach for combining the FS
@@
 
-%\section*{Acknowledgments}
+\section*{Acknowledgments}
+
+This work is partially funded by the ERDF (European Regional
+Development Fund) through the COMPETE Programme and by FCT (Portuguese
+Foundation for Science and Technology) within projects SIBILA
+(NORTE-07-0124-FEDER-000059) and PEst
+(FCOMP-01-0124-FEDER-037281). Miguel Areias is funded by the FCT grant
+SFRH/BD/69673/2010.
 

### 2015-05-26 paper.tex
paper -> revising 1

@@
 
-We have presented a simple and novel approach for combining the FS
-design with batched scheduling. Experimental results showed that on
-average, the extra structures required for the combination do not seem
-to have a big impact in the performance. Further work will include
-experimentation in other real world applications.
+Local and batched scheduling are arguably two of the most well-known
+tabling scheduling techniques. The major difference between them is
+that local propagates answers only after all answer are found, while
+batched propagates answers immediately after they are found. Batched
+scheduling is a useful strategy in tabled logic programs that require
+an eager propagation of answers and/or do not require the complete set
+of answers to be found (partial evaluation of the search space).
+
+In this work, we have presented the PAC strategy, which is a simple
+and novel approach for combining the FS design with batched
+scheduling. PAC splits answer representation from answer propagation,
+allowing the first to be publicly shared among threads while the
+second is private to each thread.
+
+Experimental results in worst-case scenarios showed that on average,
+the PAC strategy does not seem to have a big impact in the
+performance, however it still leaves room for further improvements
+specially in the extra structures required to control the propagated
+answers. Further work will include the usage of time-stamped tries to
+minimize the search for the propagated answers and new real-world
+problems that will allow us to improve and consolidate our framework.
 
@@
 
-This work is partially funded by the ERDF (European Regional
-Development Fund) through the COMPETE Programme and by FCT (Portuguese
-Foundation for Science and Technology) within projects SIBILA
-(NORTE-07-0124-FEDER-000059) and PEst
-(FCOMP-01-0124-FEDER-037281). Miguel Areias is funded by the FCT grant
-SFRH/BD/69673/2010.
+%% This work is partially funded by the ERDF (European Regional
+%% Development Fund) through the COMPETE Programme and by FCT (Portuguese
+%% Foundation for Science and Technology) within projects SIBILA
+%% (NORTE-07-0124-FEDER-000059) and PEst
+%% (FCOMP-01-0124-FEDER-037281). Miguel Areias is funded by the FCT grant
+%% SFRH/BD/69673/2010.
 

### 2015-05-26 paper.tex
paper -> revising 1

@@
 are only returned when all program clauses for the subgoal at hand
-were resolved. Batched scheduling favors forward execution first,
-backtracking next, and consuming answers or completion last. It thus
-tries to delay the need to move around the search tree by batching the
-return of answers to repeated subgoals. When new answers are found for
-a particular tabled subgoal, they are added to the table space and the
-evaluation continues.
+were resolved, i.e. after the complete evaluation of subgoal's search
+space. Local scheduling has the advantage of minimizing the size of
+clusters of dependent subgoals, however it delays propagation of
+answers and requires the complete evaluation of the search
+space.
+
+Batched scheduling favors forward execution first, backtracking next,
+and consuming answers or completion last. It thus tries to delay the
+need to move around the search tree by batching the return of answers
+to repeated subgoals. When new answers are found for a particular
+tabled subgoal, they are added to the table space and the evaluation
+continues. Batched scheduling can be an useful strategy in tabled
+logic programs that require an eager propagation of answers and/or do
+not require the complete set of answers to be found (partial
+evaluation of the search space).
 
@@
 dependent subgoals in a CDS are completely
-evaluated~\cite{Sagonas-98}.
+evaluated~\cite{Sagonas-98}. Batched scheduling can be an useful
+strategy in tabled logic programs that require an eager propagation of
+answers and/or do not require the complete set of answers to be found
+(partial evaluation of the search space).
 
@@
 propagated to the thread's repeated calls.
+
+\begin{figure}[!ht]
+%\vspace{-\intextsep}
+\centering
+\includegraphics[width=10.5cm]{figures/pcc_key-idea.pdf}
+\caption{PAC splits answer propagation and answer representation}
+\label{fig_tabtries_pcc_key-idea}
+%\vspace{-1.0\intextsep}
+\end{figure}
 
@@
 scheduling. PAC splits answer representation from answer propagation,
-allowing the first to be publicly shared among threads while the
-second is private to each thread.
+and allows the first to be publicly shared among threads while the
+second to be private to each thread.
 

### 2015-05-26 paper.tex
paper -> revising 1

@@
 that new answers for a given subgoal call are correctly propagated one
-and only once to all thread’s repeated calls. To solve that, we propose
-an extension to the table space data structures, which we named
-\emph{Private Answer Chaining} (PAC), as a way to keep track, per
-thread and subgoal call, of the answers that were already found and
-propagated to the thread's repeated calls.
+and only once to all thread’s repeated calls. In the FS design answer
+representation and answer propagation is stored in the answer trie
+data structure, thus threads are unable to know if they have or not
+have propagated an answer whenever it appears during the
+evaluation. To solve that, we propose an extension to the table space
+data structures, which we named \emph{Private Answer Chaining} (PAC),
+as a way to keep track, per thread and subgoal call, of the answers
+that were already found and propagated to the thread's repeated
+calls. Figure~\ref{fig_tabtries_pcc_key-idea} illustrates the key idea
+of the PAC strategy. In a nutshell, PAC splits answer representation
+from answer propagation, and allows the first to be kept publicly
+shared among threads in the answer trie data structure, and the second
+to be privately stored in the subgoal frame data structure of each
+thread. Thus, with PAC, whenever a answer appears during the
+evaluation, a thread knows if the answer was already propagated or not
+propagated, just by consulting its own subgoal frame data structure.
 
@@
 \includegraphics[width=10.5cm]{figures/pcc_key-idea.pdf}
-\caption{PAC splits answer propagation and answer representation}
+\caption{PAC splits answer representation and answer propagation}
 \label{fig_tabtries_pcc_key-idea}

### 2015-05-26 paper.tex
paper -> revising 1

@@
 
-\begin{figure}[!ht]
-%\vspace{-\intextsep}
+\begin{wrapfigure}{r}{6.5cm}
+\vspace{-\intextsep}
 \centering
-\includegraphics[width=10.5cm]{figures/pcc_key-idea.pdf}
+\includegraphics[width=6.5cm]{figures/pcc_key-idea.pdf}
 \caption{PAC splits answer representation and answer propagation}
 \label{fig_tabtries_pcc_key-idea}
-%\vspace{-1.0\intextsep}
-\end{figure}
+\vspace{-0.5\intextsep}
+\end{wrapfigure}
+
+%% \begin{figure}[!ht]
+%% \centering
+%% \includegraphics[width=7.5cm]{figures/pcc_key-idea.pdf}
+%% \caption{PAC splits answer representation and answer propagation}
+%% \label{fig_tabtries_pcc_key-idea}
+%% \end{figure}
 
@@
 
-We now present experimental results about the usage of the FS design
-with batched scheduling. To put results in perspective, we will be
-showing also the results for local scheduling and for the NS
-design.
+We now present experimental results about the usage of PAC in the FS
+design with batched scheduling. Since without PAC the FS design would
+not be able to be used with batched scheduling, to put results in
+perspective, we will be showing also the results for local scheduling
+and for the NS design.
 
@@
 in~\cite{Areias-12b}. These benchmarks create \emph{worst case
-  scenarios}, where we are able to show the lowest bounds in terms of
-performance that each design might achieve when applied/used with
-other real world applications/programs.
+  scenarios}, where we are able to show the lowest bounds of
+performance that each design might achieve when applied/used in other
+real world applications/programs.
 
@@
 & \multicolumn{1}{|c}{\bf Local}
-& \multicolumn{1}{c}{\bf Batched}\\
+& \multicolumn{1}{c}{\bf Batched (PAC)}\\
 \hline\hline

### 2015-05-26 paper.tex
paper -> revising 1

@@
 design. For the average and maximum overhead ratios, in the FS design,
-the best strategy is always the local scheduling.
+the best strategy is always the local scheduling. Additionally, the
+difference between local and batched scheduling in the FS is higher
+than in the NS design, which indicates the overhead that the PAC
+strategy brings to the FS design. Recall, whenever an answer is found
+during the evaluation, PAC requires threads to traverse their private
+consumer data structures to check if the answer was already
+propagated. This is the reason for overhead that we are observing in
+the experimental results.
 
@@
 strategies is quite close. For the maximum overhead ratio, the
-difference between both strategies is significantly higher.
+difference between both strategies is significantly higher than in the
+NS design.
 
-%% As we scale the number of threads, one can observe that,
-%% for the NS design both scheduling strategies have similar minimum,
-%% average and maximum overhead ratios. For the FS design, the best
-%% minimum overhead ratio on $1$, $8$, $16$, $24$ and $32$ threads is
-%% always the batched scheduling. For the average and maximum overhead
-%% ratio, local scheduling is always better than batched
-%% scheduling. However, for the average ratio, the results between both
-%% strategies is quite close. We consider this to be quite significant,
-%% once the FS design with batched scheduling requires the support of the
-%% PAC procedure. For the maximum overhead ratio, the difference between
-%% both strategies is higher than for the average overhead ratio, however
-%% we hope to reduce the difference with further improvements in the PAC
-%% procedure.
+In resume, the experimental results show that on average, the PAC
+strategy does not seem to have a big impact in the performance,
+however it still leaves room for further improvements, because the
+difference between local and batched scheduling is higher in the FS
+design than in the NS design.
 

### 2015-05-26 paper.tex
paper -> revising 1

@@
 most successful tabling scheduling strategies. In this work, we
-propose an extension of the Full-Sharing design to support batched
-scheduling. Experimental results show that, despite the extra data
-structures required to support the extension, the FS design remains
-quite competitive.\\
+propose the PAC strategy to combine the Full-Sharing design with
+batched scheduling. Experimental results show that, PAC is a good
+first approach, since with it the FS design remains quite
+competitive.\\
 
@@
 logic programs that require an eager propagation of answers and/or do
-not require the complete set of answers to be found (partial
-evaluation of the search space).
+not require the complete set of answers to be found.
 
@@
 \begin{wrapfigure}{r}{4.5cm}
-%\vspace{-\intextsep}
+\vspace{-\intextsep}
 \centering
@@
 \label{fig_table_space}
-%\vspace{-0.5\intextsep}
+\vspace{-0.5\intextsep}
 \end{wrapfigure}
@@
 computation, to a \emph{Full-Sharing} (FS) design, where threads share
-the complete table space. Figure~\ref{fig_table_space} shows Yap's
+the complete table space. Figure~\ref{fig_yap_mt_support} shows Yap's
 multithreaded table space organization for the NS and FS designs,
@@
 perspective, we will be showing also the results for local scheduling
-and for the NS design.
-
-The environment for our experiments was a machine with 32-Core AMD
-Opteron (TM) Processor 6274 (2 sockets with 16 cores each) with 32G of
-main memory, running the Linux kernel is the 3.16.7-200.fc20.x86\_64
-with Yap Prolog 6.3. For the experiments, we used the \emph{TabMalloc}
-memory allocator~\cite{Areias-12b} and the set of benchmarks described
-in~\cite{Areias-12b}. These benchmarks create \emph{worst case
-  scenarios}, where we are able to show the lowest bounds of
-performance that each design might achieve when applied/used in other
-real world applications/programs.
+and for the NS design. The environment for our experiments was a
+machine with 32-Core AMD Opteron (TM) Processor 6274 (2 sockets with
+16 cores each) with 32G of main memory, running the Linux kernel is
+the 3.16.7-200.fc20.x86\_64 with Yap Prolog 6.3. For the experiments,
+we used the \emph{TabMalloc} memory allocator~\cite{Areias-12b} and
+the set of benchmarks described in~\cite{Areias-12b}. These benchmarks
+create \emph{worst case scenarios}, where we are able to show the
+lowest bounds of performance that each design might achieve when
+applied/used in other real world applications/programs.
 
@@
 consumer data structures to check if the answer was already
-propagated. This is the reason for overhead that we are observing in
-the experimental results.
-
-As we scale the number of threads, for the NS design both scheduling
-strategies have similar minimum, average and maximum overhead
-ratios. For the FS design, the best minimum overhead ratio is always
-for the batched scheduling. For the average and maximum overhead
-ratio, local scheduling is always better than batched
-scheduling. However, for the average ratio, the results between both
-strategies is quite close. For the maximum overhead ratio, the
-difference between both strategies is significantly higher than in the
-NS design.
-
-In resume, the experimental results show that on average, the PAC
-strategy does not seem to have a big impact in the performance,
-however it still leaves room for further improvements, because the
-difference between local and batched scheduling is higher in the FS
-design than in the NS design.
+propagated. This is the reason for difference in the overheads that we
+are observing in the experimental results. As we scale the number of
+threads, for the NS design both scheduling strategies have similar
+minimum, average and maximum overhead ratios. For the FS design, the
+best minimum overhead ratio is always for the batched scheduling. For
+the average and maximum overhead ratio, local scheduling is always
+better than batched scheduling. However, for the average ratio, the
+results between both strategies is quite close. For the maximum
+overhead ratio, the difference between both strategies is
+significantly higher than in the NS design. In conclusion,
+experimental results show that on average, the PAC strategy does not
+seem to have a big impact in the performance, however it still leaves
+room for further improvements. The difference between both local and
+batched scheduling is significantly higher in the FS design than in
+the NS design.
 
@@
 an eager propagation of answers and/or do not require the complete set
-of answers to be found (partial evaluation of the search space).
-
-In this work, we have presented the PAC strategy, which is a simple
-and novel approach for combining the FS design with batched
+of answers to be found (partial evaluation of the search space). In
+this work, we have presented the PAC strategy, which is a simple and
+novel approach for combining the FS design with batched
 scheduling. PAC splits answer representation from answer propagation,
@@
 
+This work is funded by the project SIBILA (NORTE-07-0124-FEDER-000059)
+under the North Portugal Regional Operational Programme (ON.2 – O Novo
+Norte) and under the National Strategic Reference Framework (NSRF),
+through the European Regional Development Fund (ERDF), and by national
+funds, through the Portuguese funding agency, Fundação para a Ciência
+e a Tecnologia (FCT).
+
 %% This work is partially funded by the ERDF (European Regional

### 2015-05-26 paper.tex
paper -> revising 1

@@
 \begin{figure}[!ht]
-%\vspace{-0.5\intextsep}
+\vspace{-0.5\intextsep}
 \centering
@@
 \label{fig_yap_mt_support}
-%\vspace{-1.0\intextsep}
+\vspace{-1.0\intextsep}
 \end{figure}
@@
 
+\begin{wrapfigure}{r}{6.5cm}
+\vspace{-\intextsep}
+\centering
+\includegraphics[width=6.5cm]{figures/pcc_key-idea.pdf}
+\caption{PAC splits answer representation and answer propagation}
+\label{fig_tabtries_pcc_key-idea}
+\vspace{-0.5\intextsep}
+\end{wrapfigure}
+
 In this section, we describe our proposal to support the combination
@@
 propagated, just by consulting its own subgoal frame data structure.
-
-\begin{wrapfigure}{r}{6.5cm}
-\vspace{-\intextsep}
-\centering
-\includegraphics[width=6.5cm]{figures/pcc_key-idea.pdf}
-\caption{PAC splits answer representation and answer propagation}
-\label{fig_tabtries_pcc_key-idea}
-\vspace{-0.5\intextsep}
-\end{wrapfigure}
 
@@
 \begin{figure}[!ht]
-%\vspace{-\intextsep}
+\vspace{-\intextsep}
 \centering
@@
 \label{fig_tabtries_pcc}
-%\vspace{-1.0\intextsep}
+\vspace{-0.5\intextsep}
 \end{figure}
@@
 and it is using a direct answer chaining to access the node
-$LN_1$. Thread $T_{k-1}$ was already found three answers for the
+$LN_1$. Thread $T_{k-2}$ was already found three answers for the
 subgoal call and it is using the hash trie mechanism within its
@@
 between themselves, thus that repeated calls belonging to thread
-$T_{k-1}$ can consume the answers as in the original mechanism.
+$T_{k-2}$ can consume the answers as in the original mechanism.
 
@@
 showing the situation where the subgoal call is completed and both
-threads $T_1$ and $T_{k-1}$ have already chained the leaf nodes inside
+threads $T_1$ and $T_{k-2}$ have already chained the leaf nodes inside
 the answer trie and removed their private chaining structures.
@@
 & \multicolumn{1}{|c}{\bf Local}
-& \multicolumn{1}{c}{\bf Batched (PAC)}\\
+& \multicolumn{1}{c}{\bf Batched}\\
 \hline\hline

### 2015-05-26 paper.tex
paper -> revising 1

@@
 some limitations of traditional Prolog systems in dealing with
-redundant sub-computations and recursion. Tabling consists of storing
+redundant sub-computations and recursion. Tabling consists in storing
 intermediate answers for subgoals in a proper data structure, called
@@
 the thread at hand in order to ensure that new answers for a given
-subgoal call are correctly propagated one and only once to all
+subgoal call are correctly propagated once and only once to all
 thread's repeated calls.
@@
 machine with 32-Core AMD Opteron (TM) Processor 6274 (2 sockets with
-16 cores each) with 32G of main memory, running the Linux kernel is
-the 3.16.7-200.fc20.x86\_64 with Yap Prolog 6.3. For the experiments,
-we used the \emph{TabMalloc} memory allocator~\cite{Areias-12b} and
-the set of benchmarks described in~\cite{Areias-12b}. These benchmarks
+16 cores each) with 32G of main memory, running the Linux kernel
+3.16.7-200.fc20.x86\_64 with Yap Prolog 6.3. For the experiments, we
+used the \emph{TabMalloc} memory allocator~\cite{Areias-12b} and the
+set of benchmarks described in~\cite{Areias-12b}. These benchmarks
 create \emph{worst case scenarios}, where we are able to show the
@@
 better than batched scheduling. However, for the average ratio, the
-results between both strategies is quite close. For the maximum
+results between both strategies are quite close. For the maximum
 overhead ratio, the difference between both strategies is

### 2015-05-27 paper.tex
paper -> revising 1

@@
 
-Tabling~\cite{Chen-96} is an implementation technique that overcomes
-some limitations of traditional Prolog systems in dealing with
-redundant sub-computations and recursion. Tabling consists in storing
+Tabling~\cite{Chen-96} is a technique that overcomes some limitations
+of traditional Prolog systems in dealing with redundant
+sub-computations and recursion. Tabling consists in storing
 intermediate answers for subgoals in a proper data structure, called
@@
 records, flags and other global non-backtrackable data are stored. For
-tabled evaluation, a thread views its tables as private but, at the
-engine level, Yap has three designs~\cite{Areias-12a} that vary from a
-\emph{No-Sharing} (NS) design, where each thread allocates fully
-private tables for each new tabled subgoal called during its
-computation, to a \emph{Full-Sharing} (FS) design, where threads share
-the complete table space.
+multithreaded tabled evaluation, a thread views its tables as private
+but, at the engine level, Yap has three
+designs~\cite{Areias-12a}. These designs vary from a \emph{No-Sharing}
+(NS) design, where each thread allocates private tables for each new
+tabled subgoal called during its evaluation, to a \emph{Full-Sharing}
+(FS) design, where threads share the complete table space.
 
@@
 exists, it is not possible to determine if the answer is new or
-repeated for a certain thread without further support. However, for
-local scheduling this is not a problem because, for repeated and new
+repeated for a particular thread without further support. For local
+scheduling this is not a problem because, for repeated and new
 answers, local scheduling always fails. The problem is with batched
-scheduling that requires further support to work with the FS design
-since, with batched scheduling, only repeated answers should fail. We
-have thus to detect when a newly found answer is also a new answer for
-the thread at hand in order to ensure that new answers for a given
-subgoal call are correctly propagated once and only once to all
-thread's repeated calls.
+scheduling that requires that only the repeated answers should
+fail. Threads have then to detect during the evaluation whether an
+answer is new in their evaluation and must be propagated or an answer
+is repeated and the evaluation should fail.
 
@@
 way to keep track, per thread and subgoal call, of the answers that
-were already found and propagated to all repeated calls. We discuss in
-detail our proposal for extending the FS design with batched
-scheduling and we present a performance analysis comparison between
-local and batched scheduling. Experimental results show that, despite
-the extra PAC data structures required to support batched scheduling
-with the FS design, the execution time of the combination is still
-quite competitive when comparing against the FS design with local
-scheduling.
+were already found and already propagated. We discuss in detail our
+proposal for extending the FS design with batched scheduling and we
+present a performance analysis comparison between local and batched
+scheduling. Experimental results show that, despite the extra PAC data
+structures required to support batched scheduling with the FS design,
+the execution time of the combination is still quite competitive when
+comparing against the FS design with local scheduling.
 
@@
 
-This section introduces some background needed for the following
-sections.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Table Space Organization}
-
 The basic idea behind tabling is straightforward: programs are
 evaluated by storing answers for tabled subgoals in an appropriate
-data space, called the \emph{table space}. Similar calls to tabled
-subgoals are not re-evaluated against the program clauses, instead
-they are resolved by consuming the answers already stored in their
-table entries. During this process, as further new answers are found,
-they are stored in their tables and later returned to all similar
-calls.
+data space, called the \emph{table space}. Similar calls\footnote{We
+  are considering \emph{variant-based tabling} similarity. Two tabled
+  subgoals A and B are variants if they can be made identical by
+  \emph{variable} renaming. For example, \emph{p(X,1,Y)} and
+  \emph{p(Y,1,Z)} are variants because both can be transformed into
+  \emph{p(VAR0,1,VAR1)}.} to tabled subgoals are not re-evaluated
+against the program clauses, instead they are resolved by consuming
+the answers already stored in their table entries. During this
+process, as further new answers are found, they are stored in their
+tables and later returned to all similar calls.
 
@@
 
-Local scheduling schedules the evaluation of a program in a
-breath-first manner. It favors backtracking first with completion
-instead of forward execution, leaving the consumption of answers for
-last. Local scheduling only allows a \emph{Cluster of Dependent
-  Subgoals} (CDS) to return answers after a fix-point has been
-reached~\cite{Freire-96}. In other words, local scheduling tries to
-keep a CDS as minimal as possible, thus creating less complex
-dependencies between subgoals, which causes a sooner completion of
-subgoals.
+Local scheduling evaluates a tabled logic program in a breath-first
+manner. It favors backtracking first with completion instead of
+forward execution, leaving the consumption of answers for last. Local
+scheduling only allows a \emph{Cluster of Dependent Subgoals} (CDS) to
+return answers after a fix-point has been reached~\cite{Freire-96}. In
+other words, local scheduling tries to keep a CDS as minimal as
+possible, thus creating less complex dependencies between subgoals,
+which causes a sooner completion of subgoals.
 
@@
 
-The algorithm receives two arguments: the newly answer found during
+The algorithm receives two arguments: the newly found answer during
 the evaluation (\emph{ANS}) and the subgoal frame which corresponds to
@@
 Table~\ref{tab_batched_overhead} shows the overhead ratios, when
-compared with the NS design with 1 thread (running with local
-scheduling without TabMalloc), for the NS and FS designs, when running
-1, 8, 16, 24 and 32 threads, with TabMalloc and, local scheduling
-(column \emph{Local}) and batched scheduling (column \emph{Batched})
+comparing against the NS design with 1 thread (running with local
+scheduling without TabMalloc), for the NS and FS designs, with 1, 8,
+16, 24 and 32 threads, using TabMalloc and, local scheduling (column
+\emph{Local}) and batched scheduling (column \emph{Batched})
 strategies. In order to give a fair weight to each benchmark, the

### 2015-05-27 paper.tex
paper -> revising 1 -> end for noew

@@
 \begin{abstract}
-Tabling is an implementation technique that overcomes some limitations
-of traditional Prolog systems in dealing with redundant
-sub-computations and recursion. Multithreading in Prolog is the
-ability to concurrently perform computations, in which each
-computation runs independently but shares the program clauses. When
-multithreading is combined with tabling, we have the best of both
-worlds, since we can exploit the combination of higher procedural
-control with higher declarative semantics. To support this
-combination, the Yap Prolog system has, at engine level, multiple
-designs that vary from a No-Sharing design, where each thread
-allocates fully private tables for each new tabled subgoal called
-during its computation, to a Full-Sharing design, where threads share
-the complete table space. Arguably, batched scheduling is one of the
-most successful tabling scheduling strategies. In this work, we
-propose the PAC strategy to combine the Full-Sharing design with
-batched scheduling. Experimental results show that, PAC is a good
+Tabling is a technique that overcomes some limitations of traditional
+Prolog systems in dealing with redundant sub-computations and
+recursion. Multithreading in Prolog is the ability to concurrently
+perform computations, in which each computation runs independently but
+shares the program clauses. When multithreading is combined with
+tabling, we have the best of both worlds, since we can exploit the
+combination of higher procedural control with higher declarative
+semantics. To support this combination, the Yap Prolog system has, at
+engine level, multiple designs that vary from a No-Sharing design,
+where each thread allocates fully private tables for each new tabled
+subgoal called during its computation, to a Full-Sharing design, where
+threads share the complete table space. Arguably, batched scheduling
+is one of the most successful tabling scheduling strategies. In this
+work, we propose the PAC strategy to combine the Full-Sharing design
+with batched scheduling. Experimental results show that, PAC is a good
 first approach, since with it the FS design remains quite
@@
 the XSB Prolog system and in particular in the SLG-WAM
-engine~\cite{Sagonas-98}, the most successful engine of XSB. The
-success of SLG-WAM led to several alternative implementations that
-differ in the execution rule, in the data-structures used to implement
-tabling, and in the changes to the underlying Prolog
-engine. Implementations of tabling are now widely available in systems
+engine~\cite{Sagonas-98}, the most successful engine of
+XSB. Implementations of tabling are now widely available in systems
 like Yap Prolog, B-Prolog, ALS-Prolog, Mercury, Ciao Prolog and more
@@
 multithreading with tabling. Yap implements a SWI-Prolog compatible
-multithreading library~\cite{Wielemaker-03}. Yap's threads have their
-own execution stacks and only share the code area where predicates,
-records, flags and other global non-backtrackable data are stored. For
-multithreaded tabled evaluation, a thread views its tables as private
-but, at the engine level, Yap has three
-designs~\cite{Areias-12a}. These designs vary from a \emph{No-Sharing}
-(NS) design, where each thread allocates private tables for each new
-tabled subgoal called during its evaluation, to a \emph{Full-Sharing}
-(FS) design, where threads share the complete table space.
+multithreading library~\cite{Wielemaker-03}. For multithreaded tabled
+evaluation, a thread views its tables as private but, at the engine
+level, Yap has three designs~\cite{Areias-12a}. These designs vary
+from a \emph{No-Sharing} (NS) design, where each thread allocates
+private tables for each new tabled subgoal called during its
+evaluation, to a \emph{Full-Sharing} (FS) design, where threads share
+the complete table space.
 
@@
 to complete subgoals as soon as possible. When new answers are found,
-they are added to the table space and the evaluation fails. Answers
-are only returned when all program clauses for the subgoal at hand
-were resolved, i.e. after the complete evaluation of subgoal's search
-space. Local scheduling has the advantage of minimizing the size of
-clusters of dependent subgoals, however it delays propagation of
-answers and requires the complete evaluation of the search
-space.
+they are added to the table space and the evaluation fails. Local
+scheduling has the advantage of minimizing the size of clusters of
+dependent subgoals, however it delays propagation of answers and
+requires the complete evaluation of the search space.
 
@@
 scheduling. Experimental results show that, despite the extra PAC data
-structures required to support batched scheduling with the FS design,
-the execution time of the combination is still quite competitive when
-comparing against the FS design with local scheduling.
+structures required, the execution time is still quite competitive.
 
 The remainder of the paper is organized as follows. First, we briefly
-introduce some background and related work. Then, we describe our
-approach for combining batched scheduling with the FS design and we
-discuss the most important implementation details. Finally, we present
-experimental results and we end by outlining some conclusions.
+introduce some background and related work. Then, we describe our PAC
+approach and we discuss the most important implementation
+details. Finally, we present experimental results and we end by
+outlining some conclusions.
 
@@
 \label{fig_yap_mt_support}
-\vspace{-1.0\intextsep}
+\vspace{-0.5\intextsep}
 \end{figure}
@@
 
-On the other hand, batched scheduling schedules the evaluation of a
-program in a depth-first manner. It favors the forward execution first
-instead of backtracking, leaving the consumption of answers and
-completion for last. It thus tries to delay the need to move around
-the search tree by batching the return of answers. When new answers
-are found for a particular tabled subgoal, they are added to the table
-space and the execution continues. For some situations, this results
-in creating dependencies to older subgoals, therefore enlarging the
-current CDS and delaying the fix-point that guarantees that all
-dependent subgoals in a CDS are completely
-evaluated~\cite{Sagonas-98}. Batched scheduling can be an useful
-strategy in tabled logic programs that require an eager propagation of
-answers and/or do not require the complete set of answers to be found
-(partial evaluation of the search space).
+On the other hand, batched scheduling evaluates a tabled logic program
+in a depth-first manner. It favors the forward execution first instead
+of backtracking, leaving the consumption of answers and completion for
+last. It thus tries to delay the need to move around the search tree
+by batching the return of answers. When new answers are found for a
+particular tabled subgoal, they are added to the table space and the
+execution continues. For some situations, this results in creating
+dependencies to older subgoals, therefore enlarging the current CDS
+and delaying the fix-point that guarantees that all dependent subgoals
+in a CDS are completely evaluated~\cite{Sagonas-98}. Batched
+scheduling can be an useful strategy in tabled logic programs that
+require an eager propagation of answers and/or do not require the
+complete set of answers to be found.
 
@@
 \begin{wrapfigure}{r}{6.5cm}
-\vspace{-\intextsep}
+\vspace{-0.5\intextsep}
 \centering
@@
 In this section, we describe our proposal to support the combination
-of batched scheduling with the FS design. Remember that batched
-scheduling requires further support to work with the FS design
-because, with batched scheduling, we have to detect when a newly found
-answer is also a new answer for the thread at hand in order to ensure
-that new answers for a given subgoal call are correctly propagated one
-and only once to all thread’s repeated calls. In the FS design answer
+of batched scheduling with the FS design. In the FS design answer
 representation and answer propagation is stored in the answer trie
@@
 to be privately stored in the subgoal frame data structure of each
-thread. Thus, with PAC, whenever a answer appears during the
-evaluation, a thread knows if the answer was already propagated or not
-propagated, just by consulting its own subgoal frame data structure.
-
-%% \begin{figure}[!ht]
-%% \centering
-%% \includegraphics[width=7.5cm]{figures/pcc_key-idea.pdf}
-%% \caption{PAC splits answer representation and answer propagation}
-%% \label{fig_tabtries_pcc_key-idea}
-%% \end{figure}
+thread.
 
@@
 
-The PAC procedure works at the subgoal frame level, which is private
-to each thread. The key idea of the PAC data structure is then to
-extend subgoal frames with an auxiliary private chaining of answers
-for each subgoal call, in order to keep track of the answers already
-found for the call. Later, if a thread completes a subgoal's
+The PAC procedure works at the subgoal frame level. The key idea of
+PAC is to extend subgoal frames with an auxiliary private chaining of
+answers for each subgoal call, in order to keep track of the answers
+already found for the call. Later, if a thread completes a subgoal's
 evaluation, i.e, if the subgoal's table is marked as complete, its PAC
@@
 Figure~\ref{fig_tabtries_pcc}(a) shows then a situation where two
-threads, $T_1$ and $T_{k-1}$, are sharing the same subgoal entry for
+threads, $T_1$ and $T_{k-2}$, are sharing the same subgoal entry for
 the call $call\_i$ when the subgoal is still under evaluation, i.e.,
@@
 seem to have a big impact in the performance, however it still leaves
-room for further improvements. The difference between both local and
-batched scheduling is significantly higher in the FS design than in
-the NS design.
+room for further improvements, since the difference between local and
+batched scheduling is higher in the FS design than in the NS design.
 
@@
 an eager propagation of answers and/or do not require the complete set
-of answers to be found (partial evaluation of the search space). In
-this work, we have presented the PAC strategy, which is a simple and
-novel approach for combining the FS design with batched
-scheduling. PAC splits answer representation from answer propagation,
-and allows the first to be publicly shared among threads while the
-second to be private to each thread.
+of answers to be found. In this work, we have presented the PAC
+strategy, which is a simple and novel approach for combining the FS
+design with batched scheduling. PAC splits answer representation from
+answer propagation, and allows the first to be publicly shared among
+threads while the second to be private to each thread.
 

### 2015-06-01 paper.tex
paper -> revising 2

@@
 \newcommand{\verbatimproperties}{\renewcommand{\baselinestretch}{0.85} \small}
-%\newcommand{\tableproperties}{\centering \big}
-\newcommand{\LINECOMMENT}[1]{$\triangleright$ #1}
-%\setlength{\belowcaptionskip}{-26pt}
 
@@
 
-\title{On Extending a Full-Sharing Multithreaded Tabling Design with
-  Batched Scheduling}
+\title{On Extending a Full-Sharing Multithreaded Tabling Design with Batched Scheduling}
 
@@
 \begin{abstract}
-Tabling is a technique that overcomes some limitations of traditional
-Prolog systems in dealing with redundant sub-computations and
-recursion. Multithreading in Prolog is the ability to concurrently
-perform computations, in which each computation runs independently but
-shares the program clauses. When multithreading is combined with
-tabling, we have the best of both worlds, since we can exploit the
-combination of higher procedural control with higher declarative
-semantics. To support this combination, the Yap Prolog system has, at
-engine level, multiple designs that vary from a No-Sharing design,
-where each thread allocates fully private tables for each new tabled
-subgoal called during its computation, to a Full-Sharing design, where
-threads share the complete table space. Arguably, batched scheduling
-is one of the most successful tabling scheduling strategies. In this
-work, we propose the PAC strategy to combine the Full-Sharing design
-with batched scheduling. Experimental results show that, PAC is a good
-first approach, since with it the FS design remains quite
-competitive.\\
+  Tabling is a technique that overcomes some limitations of
+  traditional Prolog systems in dealing with redundant
+  sub-computations and recursion. When tabling is combined with
+  multithreading, we have the best of both worlds, since we can
+  exploit the combination of higher declarative semantics with higher
+  procedural control. To support this combination, the Yap Prolog
+  system has, at engine level, multiple designs that vary from a
+  No-Sharing design, where each thread allocates fully private tables,
+  to a Full-Sharing (FS) design, where threads share the complete
+  table space. In this work, we propose an extension to the table
+  space data structures, which we named \emph{Private Answer Chaining
+    (PAC)}, as way to support batched scheduling evaluation with the
+  FS design. Batched scheduling is one of the most successful tabling
+  scheduling strategies, known to be useful when a tabled logic
+  program requires an eager propagation of answers and/or do not
+  requires the complete set of answers to be found. Experimental
+  results show that PAC is a good first approach, since with it the FS
+  design remains quite competitive.\\
 
@@
 are the only Prolog systems that support the combination of
-multithreading with tabling. Yap implements a SWI-Prolog compatible
-multithreading library~\cite{Wielemaker-03}. For multithreaded tabled
-evaluation, a thread views its tables as private but, at the engine
-level, Yap has three designs~\cite{Areias-12a}. These designs vary
-from a \emph{No-Sharing} (NS) design, where each thread allocates
-private tables for each new tabled subgoal called during its
-evaluation, to a \emph{Full-Sharing} (FS) design, where threads share
-the complete table space.
+multithreading with tabling. In this work, we will focus on Yap's
+implementation that follows a SWI-Prolog compatible multithreading
+library~\cite{Wielemaker-03}. For tabled evaluation, a thread views
+its tables as private but, at the engine level, Yap has three
+designs~\cite{Areias-12a}, which vary from a \emph{No-Sharing} (NS)
+design, where each thread allocates private tables for each new tabled
+subgoal call, to a \emph{Full-Sharing} (FS) design, where threads
+share the complete table space.
 
@@
 repeated for a particular thread without further support. For local
-scheduling this is not a problem because, for repeated and new
-answers, local scheduling always fails. The problem is with batched
-scheduling that requires that only the repeated answers should
-fail. Threads have then to detect during the evaluation whether an
-answer is new in their evaluation and must be propagated or an answer
-is repeated and the evaluation should fail.
+scheduling, this is not a problem since, for repeated and new answers,
+local scheduling always fails. The problem is with batched scheduling
+that requires that only the repeated answers should fail. Threads have
+then to detect, during batched evaluation, whether an answer is new
+and must be propagated or whether an answer is repeated and the
+evaluation should fail.
 
@@
 way to keep track, per thread and subgoal call, of the answers that
-were already found and already propagated. We discuss in detail our
-proposal for extending the FS design with batched scheduling and we
-present a performance analysis comparison between local and batched
+were already found and propagated. We discuss in detail our proposal
+for extending the FS design with batched scheduling and we present a
+performance analysis comparison between local and batched
 scheduling. Experimental results show that, despite the extra PAC data
-structures required, the execution time is still quite competitive.
+structures required to support batched scheduling with the FS design,
+the execution time of the combination is still quite competitive.
 
@@
 evaluated by storing answers for tabled subgoals in an appropriate
-data space, called the \emph{table space}. Similar calls\footnote{We
-  are considering \emph{variant-based tabling} similarity. Two tabled
-  subgoals A and B are variants if they can be made identical by
-  \emph{variable} renaming. For example, \emph{p(X,1,Y)} and
-  \emph{p(Y,1,Z)} are variants because both can be transformed into
-  \emph{p(VAR0,1,VAR1)}.} to tabled subgoals are not re-evaluated
-against the program clauses, instead they are resolved by consuming
-the answers already stored in their table entries. During this
-process, as further new answers are found, they are stored in their
-tables and later returned to all similar calls.
+data space, called the \emph{table space}. Repeated calls\footnote{We
+  are considering \emph{variant-based
+    tabling}~\cite{RamakrishnanIV-99}. Two tabled subgoals A and B are
+  variants if they can be made identical by variable renaming. For
+  example, \emph{p(X,1,Y)} and \emph{p(Y,1,Z)} are variants because
+  both can be transformed into \emph{p(VAR$_0$,1,VAR$_1$)}.} to tabled
+subgoals are not re-evaluated against the program clauses, instead
+they are resolved by consuming the answers already stored in their
+table entries. During this process, as further new answers are found,
+they are stored in their tables and later returned to all repeated
+calls.
 
@@
 \centering
-\includegraphics[width=4.5cm]{figures/table-space.pdf}
+\includegraphics[width=4.5cm]{figures/table_space.pdf}
 \caption{Table space organization}
 \label{fig_table_space}
-\vspace{-0.5\intextsep}
+\vspace{-\intextsep}
 \end{wrapfigure}
 
-Figure~\ref{fig_table_space} shows the table space organization of the
-Yap system. At the entry point we have the \emph{table entry} data
-structure. This structure is allocated when a tabled predicate is
-being compiled, so that a pointer to the table entry can be included
-in its compiled code. This guarantees that further calls to the
-predicate will access the table space starting from the same
-point. Below the table entry, we have the \emph{subgoal trie
-  structure}. Each different tabled subgoal call to the predicate at
-hand corresponds to a unique path through the subgoal trie structure,
-always starting from the table entry, passing by several subgoal trie
-data units, the \emph{subgoal trie nodes}, and reaching a leaf data
-structure, the \emph{subgoal frame}. The subgoal frame stores
-additional information about the subgoal and acts like an entry point
-to the \emph{answer trie structure}. Each unique path through the
-answer trie data units, the \emph{answer trie nodes}, corresponds to a
-different answer to the entry subgoal.
+Figure~\ref{fig_table_space} shows Yap's table space organization. At
+the entry point we have the \emph{table entry} data structure. This
+structure is allocated when a tabled predicate is being compiled, so
+that a pointer to the table entry can be included in its compiled
+code. This guarantees that further calls to the predicate will access
+the table space starting from the same point. Below the table entry,
+we have the \emph{subgoal trie structure}. Each different tabled
+subgoal call to the predicate at hand corresponds to a unique path
+through the subgoal trie structure, always starting from the table
+entry, passing by several subgoal trie data units, the \emph{subgoal
+  trie nodes}, and reaching a leaf data structure, the \emph{subgoal
+  frame}. The subgoal frame stores additional information about the
+subgoal and acts like an entry point to the \emph{answer trie
+  structure}. Each unique path through the answer trie data units, the
+\emph{answer trie nodes}, corresponds to a different answer to the
+entry subgoal.
 
@@
 from a \emph{No-Sharing} (NS) design, where each thread allocates
-fully private tables for each new tabled subgoal called during its
-computation, to a \emph{Full-Sharing} (FS) design, where threads share
-the complete table space. Figure~\ref{fig_yap_mt_support} shows Yap's
-multithreaded table space organization for the NS and FS designs,
-where an interface layer abstracts the design being used at the engine
-level. The figure illustrates the main differences between the two
-designs for a situation where several threads are evaluating the same
-tabled subgoal call $call\_i$.
+fully private tables, to a \emph{Full-Sharing} (FS) design, where
+threads share the complete table
+space. Figure~\ref{fig_yap_mt_support} shows Yap's multithreaded table
+space organization for the NS and FS designs, where an interface layer
+abstracts the design being used at the engine level. The figure
+illustrates the main differences between the two designs for a
+situation where several threads are evaluating the same tabled subgoal
+call $call\_i$.
 
 \begin{figure}[!ht]
-\vspace{-0.5\intextsep}
+%\vspace{-\intextsep}
 \centering
-\includegraphics[width=9.5cm]{figures/yap-mt.pdf}
+\includegraphics[width=9.5cm]{figures/yap_mt.pdf}
 \caption{Yap's multithreaded table space organization for the NS and FS designs}
 \label{fig_yap_mt_support}
-\vspace{-0.5\intextsep}
+\vspace{-\bigskipamount}
 \end{figure}
 
-When using the NS design, each thread allocates fully private tables
-for each new subgoal called during its computation. One can observe
-that the table entry data structure still stores the common
-information for the predicate (such as the arity or the evaluation
-strategy), and then each thread $t$ has its own cell $T_t$ inside a
-\emph{bucket array} which points to the private data structures.
+When using the NS design, one can observe that the table entry data
+structure still stores the common information for the predicate (such
+as the arity or the scheduling strategy), and then each thread $t$ has
+its own cell $T_t$ inside a \emph{bucket array} which points to the
+private data structures.
 
@@
 \begin{wrapfigure}{r}{6.5cm}
-\vspace{-0.5\intextsep}
+\vspace{-\intextsep}
 \centering
-\includegraphics[width=6.5cm]{figures/pcc_key-idea.pdf}
-\caption{PAC splits answer representation and answer propagation}
-\label{fig_tabtries_pcc_key-idea}
-\vspace{-0.5\intextsep}
+\includegraphics[width=6.4cm]{figures/pac_overview.pdf}
+\caption{PAC overview}
+\label{fig_pac_overview}
+\vspace{-\intextsep}
 \end{wrapfigure}
@@
 In this section, we describe our proposal to support the combination
-of batched scheduling with the FS design. In the FS design answer
-representation and answer propagation is stored in the answer trie
-data structure, thus threads are unable to know if they have or not
-have propagated an answer whenever it appears during the
-evaluation. To solve that, we propose an extension to the table space
-data structures, which we named \emph{Private Answer Chaining} (PAC),
-as a way to keep track, per thread and subgoal call, of the answers
-that were already found and propagated to the thread's repeated
-calls. Figure~\ref{fig_tabtries_pcc_key-idea} illustrates the key idea
-of the PAC strategy. In a nutshell, PAC splits answer representation
-from answer propagation, and allows the first to be kept publicly
-shared among threads in the answer trie data structure, and the second
-to be privately stored in the subgoal frame data structure of each
-thread.
+of batched scheduling with the FS design. In the original FS design,
+answer propagation and answer representation are both stored in the
+answer trie data structure, thus threads are unable to distinguish
+whether they have or not have propagated an answer already stored in
+the table space. To solve that, we propose an extension to the table
+space data structures, which we named \emph{Private Answer Chaining}
+(PAC), as a way to keep track, per thread and subgoal call, of the
+answers that were already found and propagated to the thread's
+repeated calls. Figure~\ref{fig_pac_overview} illustrates PAC's key
+idea. In a nutshell, PAC splits answer propagation from answer
+representation, and allows the first to be privately stored in the
+subgoal frame data structure of each thread, and the second to be kept
+publicly shared among threads in the answer trie data structure.
 
@@
 
-The PAC procedure works at the subgoal frame level. The key idea of
-PAC is to extend subgoal frames with an auxiliary private chaining of
-answers for each subgoal call, in order to keep track of the answers
-already found for the call. Later, if a thread completes a subgoal's
+The PAC procedure works at the subgoal frame level. The key idea is to
+extend subgoal frames with an auxiliary private chaining of answers
+for each subgoal call, in order to keep track of the answers already
+found for the call. Later, if a thread completes a subgoal's
 evaluation, i.e, if the subgoal's table is marked as complete, its PAC
 is made public, so that from that point on all threads can use that
-chain in complete (only reading) mode. Figure~\ref{fig_tabtries_pcc}
+chain in complete (only reading) mode. Figure~\ref{fig_pac_details}
 illustrates the new data structures involved in the implementation of
-our PAC's proposal for a situation where several threads are
+our PAC's proposal for a situation where different threads are
 evaluating the same tabled subgoal call $call\_i$.
@@
 \begin{figure}[!ht]
-\vspace{-\intextsep}
+%\vspace{-\intextsep}
 \centering
-\includegraphics[width=10.5cm]{figures/pcc.pdf}
+\includegraphics[width=10.5cm]{figures/pac.pdf}
 \caption{PAC's data structures for (a) private and (b) public
   chaining}
-\label{fig_tabtries_pcc}
-\vspace{-0.5\intextsep}
+\label{fig_pac_details}
+\vspace{-\bigskipamount}
 \end{figure}
 
-Figure~\ref{fig_tabtries_pcc}(a) shows then a situation where two
-threads, $T_1$ and $T_{k-2}$, are sharing the same subgoal entry for
-the call $call\_i$ when the subgoal is still under evaluation, i.e.,
-the subgoal is not yet complete. The current state of the evaluation
-shows an answer trie with 3 answers found for the subgoal call
-$call\_i$. For the sake of simplicity, we are omitting the internal
-answer trie nodes and we are only showing the leaf nodes nodes $LN_1$,
-$LN_2$ and $LN_3$, in the figure.
+Figure~\ref{fig_pac_details}(a) shows then a situation where two
+threads, $T_1$ and $T_{k-2}$, are sharing the same subgoal entry for a
+call $call\_i$ still under evaluation, i.e., still not yet
+completed. The current state of the evaluation shows an answer trie
+with 3 answers found for $call\_i$. For the sake of simplicity, we are
+omitting the internal answer trie nodes and we are only showing the
+leaf nodes $LN_1$, $LN_2$ and $LN_3$ of each answer.
 
@@
 structure; and (ii) a next pointer to chain the nodes in the private
-chaining. To maintain a good performance, when the number of answer
+chaining. To maintain good performance, when the number of answer
 nodes exceeds a certain threshold, we use a hash trie mechanism design
-similar to the one presented in the
-work~\cite{Areias-ijpp15}. However, since this mechanism is private to
-each thread, it does not require any concurrency support.
+similar to the one presented in~\cite{Areias-ijpp15}, but without
+concurrency support, since this mechanism is private to each thread.
 
-Going back to Fig.~\ref{fig_tabtries_pcc}(a), the PAC's data
-structures represent then two different situations where threads can
-be evaluating a subgoal call. Thread $T_1$ has only found one answer
-and it is using a direct answer chaining to access the node
-$LN_1$. Thread $T_{k-2}$ was already found three answers for the
-subgoal call and it is using the hash trie mechanism within its
-private chaining. Please note that the answer nodes are still chained
-between themselves, thus that repeated calls belonging to thread
-$T_{k-2}$ can consume the answers as in the original mechanism.
+PAC's data structures in Fig.~\ref{fig_pac_details}(a) represent then
+two different situations. Thread $T_1$ has only found one answer and
+it is using a direct answer chaining to access the leaf node
+$LN_1$. Thread $T_{k-2}$ was already found three answers for $call\_i$
+and it is using the hash trie mechanism within its private
+chaining. In the hash trie mechanism, the answer nodes are still
+chained between themselves, thus that repeated calls belonging to
+thread $T_{k-2}$ can consume the answers as in the original mechanism.
 
-Figure~\ref{fig_tabtries_pcc}(b) shows the state of the subgoal call
+Figure~\ref{fig_pac_details}(b) shows the state of the subgoal call
 after completion. When a thread $T$ completes a subgoal call, it frees
@@
 no other thread has done that, then thread $T$ not only follows its
-private chaining mechanism as it would for freeing its private nodes,
-but also, follows the pointers to the answer trie leaf nodes in order
+private chaining mechanism, as it would for freeing its private nodes,
+but also follows the pointers to the answer trie leaf nodes in order
 to create a chain inside the answer trie. Since this procedure is done
 inside a critical region, no more than one thread can be doing this
-chaining process. Thus, in Fig.~\ref{fig_tabtries_pcc}(b), we are
-showing the situation where the subgoal call is completed and both
-threads $T_1$ and $T_{k-2}$ have already chained the leaf nodes inside
-the answer trie and removed their private chaining structures.
+chaining process. Thus, in Fig.~\ref{fig_pac_details}(b), we are
+showing the situation where the subgoal call $call\_i$ is completed
+and both threads $T_1$ and $T_{k-2}$ have already chained the leaf
+nodes inside the answer trie and removed their private chaining
+structures.
 
@@
 
-At the tabling engine, the major difference between local and batched
-scheduling is in the tabling operation \emph{tabled new answer}, where
-we decide what to do when an answer is found during the
+The major difference between local and batched scheduling, at the
+engine level, is in the \emph{tabled new answer} operation, where we
+decide what to do when a new answer is found during the
 evaluation. This operation checks whether a newly found answer is
 already in the corresponding answer trie structure and, if not,
-inserts it. Algorithm~\ref{alg_table_new_answer_batched} shows how we
+inserts it. For local scheduling, it then fails and, for batched
+scheduling, it proceeds with forward
+execution. Algorithm~\ref{alg_table_new_answer_batched} shows how we
 have extended this operation to support the FS design with batched
@@
 
+\vspace{-\bigskipamount}
 \begin{algorithm} [!ht]
 \caption{tabled\_new\_answer(answer ANS, subgoal frame SF)}
-\algsetup{indent=0.65cm}
 \begin{algorithmic}[1]
@@
   \STATE $chain \gets check\_insert\_consumer\_chain(leaf, SF)$
-  \IF {$is\_answer\_marked\_as\_found(chain) = True$}
+  \IF {$is\_answer\_marked\_as\_found(chain)$}
     \RETURN $failure$
@@
 \end{algorithm}
+\vspace{-\bigskipamount}
 
@@
 answer \emph{ANS} has found (line 6). At the end (lines 7 to 10), it
-returns failure if local scheduling is active (line 8), otherwise,
+returns failure, if local scheduling is active (line 8), otherwise,
 batched scheduling is active, and it proceeds by propagating the
@@
 design with batched scheduling. Since without PAC the FS design would
-not be able to be used with batched scheduling, to put results in
-perspective, we will be showing also the results for local scheduling
-and for the NS design. The environment for our experiments was a
-machine with 32-Core AMD Opteron (TM) Processor 6274 (2 sockets with
-16 cores each) with 32G of main memory, running the Linux kernel
+not be able to be used with batched scheduling, to put PAC's results
+in perspective, we will be showing also the results for local
+scheduling and for the NS design. The environment for our experiments
+was a machine with 32-Core AMD Opteron (TM) Processor 6274 (2 sockets
+with 16 cores each) with 32GB of main memory, running the Linux kernel
 3.16.7-200.fc20.x86\_64 with Yap Prolog 6.3. For the experiments, we
@@
 comparing against the NS design with 1 thread (running with local
-scheduling without TabMalloc), for the NS and FS designs, with 1, 8,
-16, 24 and 32 threads, using TabMalloc and, local scheduling (column
-\emph{Local}) and batched scheduling (column \emph{Batched})
-strategies. In order to give a fair weight to each benchmark, the
-overhead ratio is calculated as follows. We begin by running ten times
+scheduling and without TabMalloc), for the NS and FS designs with 1,
+8, 16, 24 and 32 threads, using local scheduling (column \emph{Local})
+and batched scheduling (column \emph{Batched}) strategies with
+TabMalloc. In order to give a fair weight to each benchmark, the
+overhead ratio is calculated as follows. We begin by running 10 times
 each benchmark $B$ for each design $D$ with $T$ threads. Then, we
@@
 to put it in perspective against the base time, which is the average
-of the ten runs of the NS design with $1$ thread ($NS_{B1}$). For
-that, we use the following formula for the overhead $O_{DBT} = D_{BT}
-/ NS_{B1}$. After calculating all the overheads $O_{DBT}$ for a
-certain design $D$ and number of threads $T$ corresponding to the
-several benchmarks $B$, we calculate the respective minimum, average,
-maximum and standard deviation overhead ratios.
+of 10 runs of the NS design with 1 thread ($NS_{B1}$). For that, we
+use the following formula for the overhead $O_{DBT} = D_{BT} /
+NS_{B1}$. After calculating all the overheads $O_{DBT}$ for a certain
+design $D$ and number of threads $T$ corresponding to the several
+benchmarks $B$, we calculate the respective minimum, average, maximum
+and standard deviation overhead ratios (rows \emph{Min}, \emph{Avg},
+\emph{Max} and \emph{StD} in Table~\ref{tab_batched_overhead}).
 
@@
 \caption{Overhead ratios, when compared with the NS design with $1$
-  thread (running with local scheduling and without TabMalloc) for the NS
-  and FS designs (with TabMalloc), when running 1, 8, 16, 24 and 32
-  threads with local and batched scheduling (best ratios by row and by
+  thread (running with local scheduling and without TabMalloc) for the
+  NS and FS designs (with TabMalloc) when running 1, 8, 16, 24 and 32
+  threads with local and batched scheduling (best ratios by row and
   design for the Minimum, Average and Maximum are in bold)}
@@
 
-By observing Table~\ref{tab_batched_overhead}, we can see that, for 1
-thread, on the minimum overhead ratio, local scheduling is better on
-the NS design, while batched scheduling is better in the FS
-design. For the average and maximum overhead ratios, in the FS design,
-the best strategy is always the local scheduling. Additionally, the
-difference between local and batched scheduling in the FS is higher
-than in the NS design, which indicates the overhead that the PAC
-strategy brings to the FS design. Recall, whenever an answer is found
-during the evaluation, PAC requires threads to traverse their private
-consumer data structures to check if the answer was already
-propagated. This is the reason for difference in the overheads that we
-are observing in the experimental results. As we scale the number of
-threads, for the NS design both scheduling strategies have similar
-minimum, average and maximum overhead ratios. For the FS design, the
-best minimum overhead ratio is always for the batched scheduling. For
-the average and maximum overhead ratio, local scheduling is always
-better than batched scheduling. However, for the average ratio, the
-results between both strategies are quite close. For the maximum
-overhead ratio, the difference between both strategies is
-significantly higher than in the NS design. In conclusion,
-experimental results show that on average, the PAC strategy does not
+By observing Table~\ref{tab_batched_overhead}, we can see that batched
+scheduling always achieves the best minimum overhead ratio in the FS
+design but, for the average and maximum overhead ratios, the best
+strategy is always local scheduling. For the average and maximum
+overhead ratios, the difference between local and batched scheduling
+in the FS design is slightly higher than in the NS design, which can
+be read as an indication of the overhead that PAC introduces into the
+FS design. Recall that whenever an answer is found during the
+evaluation, PAC requires that threads traverse their private consumer
+data structures to check if the answer was already found (and propagated).
+
+As we increase the number of threads, for the NS design, both
+scheduling strategies show very close minimum, average and maximum
+overhead ratios. For the FS design, the differences are slightly
+higher. However, for the average overhead ratio, the results between
+both strategies are quite close, with batched scheduling being around
+10\% slower than local scheduling for the FS design. In summary, our
+experimental results show that, on average, the PAC strategy does not
 seem to have a big impact in the performance, however it still leaves
@@
 Local and batched scheduling are arguably two of the most well-known
-tabling scheduling techniques. The major difference between them is
-that local propagates answers only after all answer are found, while
-batched propagates answers immediately after they are found. Batched
-scheduling is a useful strategy in tabled logic programs that require
-an eager propagation of answers and/or do not require the complete set
-of answers to be found. In this work, we have presented the PAC
-strategy, which is a simple and novel approach for combining the FS
-design with batched scheduling. PAC splits answer representation from
-answer propagation, and allows the first to be publicly shared among
-threads while the second to be private to each thread.
+tabling scheduling strategies. The major difference between both is
+that local scheduling propagates answers only after all answers are
+found, while batched scheduling propagates answers immediately after
+they are found. Batched scheduling is a useful strategy in tabled
+logic programs that require an eager propagation of answers and/or do
+not require the complete set of answers to be found. In this work, we
+have presented the PAC strategy, which is a simple and novel approach
+for combining the FS design with batched scheduling. PAC splits answer
+representation from answer propagation, and allows the first to be
+publicly shared among threads while the second to be private to each
+thread.
 
-Experimental results in worst-case scenarios showed that on average,
+Experimental results in worst-case scenarios showed that, on average,
 the PAC strategy does not seem to have a big impact in the
@@
 
-This work is funded by the project SIBILA (NORTE-07-0124-FEDER-000059)
-under the North Portugal Regional Operational Programme (ON.2 – O Novo
-Norte) and under the National Strategic Reference Framework (NSRF),
-through the European Regional Development Fund (ERDF), and by national
-funds, through the Portuguese funding agency, Fundação para a Ciência
-e a Tecnologia (FCT).
-
-%% This work is partially funded by the ERDF (European Regional
-%% Development Fund) through the COMPETE Programme and by FCT (Portuguese
-%% Foundation for Science and Technology) within projects SIBILA
-%% (NORTE-07-0124-FEDER-000059) and PEst
-%% (FCOMP-01-0124-FEDER-037281). Miguel Areias is funded by the FCT grant
-%% SFRH/BD/69673/2010.
+This work is partially funded by project SIBILA
+(NORTE-07-0124-FEDER-000059) under the North Portugal Regional
+Operational Programme (ON.2 – O Novo Norte) and under the National
+Strategic Reference Framework (NSRF), through the European Regional
+Development Fund (ERDF) and, by national funds, through the Portuguese
+Foundation for Science and Technology (FCT).
 

### 2015-06-01 paper.tex
paper -> revising (end)

@@
 
-This work is partially funded by project SIBILA
-(NORTE-07-0124-FEDER-000059) under the North Portugal Regional
-Operational Programme (ON.2 – O Novo Norte) and under the National
+This work is partially funded by the North Portugal Regional
+Operational Programme (ON.2 – O Novo Norte) and by the National
 Strategic Reference Framework (NSRF), through the European Regional
-Development Fund (ERDF) and, by national funds, through the Portuguese
-Foundation for Science and Technology (FCT).
+Development Fund (ERDF) and the Portuguese Foundation for Science and
+Technology (FCT), within projects NORTE-07-0124-FEDER-000059 and
+UID/EEA/50014/2013.
 
